---
title: "Project Multivariate data Analysis"
author:
  - "Sara Vasques"
  - "Rita Pinto"
  - "Juan Pablo Martinez Aldana"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
---
<div style="text-align: justify;">

```{r}
# 2. Create folder if it doesn't exist
out_dir <- file.path("./plots")   # change base path if needed

```


# Abstract

Air pollution in urban environments is a significant public health concern.
Some evidence indicates that air pollution exposure has way more consequences than the ones that we thought previously. In terms of adverse health impacts like the reduction in life expectancy, and hospital admissions, birth outcomes, and asthma. Nevertheless, these effects exist in both economically developing and developed countries [1].

Classical PCA can be expressed as a projection-based approach, since it's a method that projects high-dimensional data into a lower-dimensional space by finding the principal components (PC) that best represent the original information. With this, we can do a dimensional reduction while preserving the essential structure of the data. [2]


Due to the reasons mentioned before, this project aims to analyze a dataset of 41 US cities to identify underlying patterns and relationships between pollution levels, climatic conditions, and demographic or industrial factors.

The study utilizes multivariate techniques, primarily **Principal Component Analysis (PCA)** and **Cluster Analysis**, to reduce the complexity of the data and group cities with similar profiles.
The objective is to translate complex statistical relationships into clear, interpretable insights.

The dataset contains the following variables:

 **so2**: Sulfur dioxide content of air in micrograms per cubic meter

 **temp**: Average annual temperature in Fahrenheit

 **manuf**: Number of manufacturing enterprises employing 20 or more workers

 **pop**: Population size (1970 census) in thousands

 **wind**: Average wind speed in miles per hour

 **precip**: Average annual precipitation in inches

 **days**: Average number of days with precipitation per year

# Preliminary analysis of the data

Before applying multivariate methods, a preliminary analysis is essential to understand the individual variables and their pairwise relationships.

```{r setup-spatial-toggle, include=FALSE}
# --- ANALYSIS CONTROL ---
# Set this to TRUE if you have a stable internet connection to generate the city map.
# If the script gets stuck on the geocoding step, set this to FALSE and re-run.
run_spatial_analysis <- FALSE
```

```{r data-loading, echo=FALSE}
# The data is loaded from the folder data/data5.csv

#data5 <- read.csv("./data/data_5.csv", header = TRUE)
data5 <- read.csv("./data/data_5.csv", header = TRUE)
head(data5)
```

```{r data-structure, echo=FALSE}
# Display the dimensions and structure of the dataset.
dim(data5)

str(data5)
```

## Numerical Data

```{r variable-selection, echo=FALSE}
# We select only the numeric variables for the analysis. (I modified the selector using seletc_if
# In order to make it more global and we can apply to any data in the future)

library(dplyr)

data5_variables <- select_if(data5, is.numeric)
head(data5_variables)
```

## Descriptive Statistics

```{r localization-measures, echo=FALSE}
# Calculate and display summary statistics for each variable.
summary(data5_variables)
```
```{r}
#graphical representation of summary values
num_data <- data5_variables[sapply(data5_variables, is.numeric)]

library(ggplot2)

for (var in names(num_data)) {

  x <- num_data[[var]]

  s <- summary(x)
  summary_df <- data.frame(
    Statistic = names(s),
    Value = as.numeric(s)
  )

  print(
    ggplot(summary_df, aes(x = Statistic, y = Value)) +
      geom_col(fill = "steelblue") +
      labs(title = paste("Summary of", var),
           x = "Statistic",
           y = "Value") +
      theme_bw()
  )
}
```

```{r}
#so2 Distribution by City
ggplot(data5, aes(x = city, y = so2)) +
  geom_boxplot(fill = "lightgreen") +
  theme_minimal() +
  labs(title = "so2 Distribution by City", x = "City", y = "so2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#correlation heatmap of the numeric variables

# install.packages(c("corrplot", "reshape2"))
library(corrplot)
library(reshape2)

corr_matrix <- cor(data5_variables, use = "complete.obs")  # handles missing values

# Basic correlation heatmap
corrplot(corr_matrix, method = "color",
         addCoef.col = "black",   # show correlation coefficients
         tl.col = "black",        # axis labels color
         number.cex = 0.8,       # font size for coefficients
         tl.cex = 0.8,            # font size for labels
         title = "Correlation Heatmap for the variables",
         mar = c(0,0,1,0))
```
```{r}
#correlation heatmap between cities (according to the variables)

library(dplyr)
library(ggplot2)
library(reshape2)

city_summary <- data5 %>%
  group_by(city) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE))

city_corr <- cor(t(city_summary %>% select(-city)))

melted_corr <- melt(city_corr)
colnames(melted_corr) <- c("City1", "City2", "Correlation")

ggplot(melted_corr, aes(x = City1, y = City2, fill = Correlation)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 2) +  # smaller text
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limit = c(0.5, 1), space = "Lab") +  # limit scale
  theme_minimal() +
  labs(title = "City-to-City Correlation Heatmap", x = "", y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size=6),  # smaller x labels
        axis.text.y = element_text(size=6),
        plot.title = element_text(hjust = 0.5, size = 14))



```
```{r}
#relationship between pop and manuf
ggplot(data5, aes(x = pop, y = manuf)) +
  geom_point(color = "tomato") +
  theme_minimal() +
  labs(title = "pop vs manuf", x = "Population", y = "manuf")
ggplot(data5, aes(x = pop, y = manuf, color = pop)) +
  geom_point(size = 3, alpha = 0.7) +
  theme_minimal()
```


We identify that `manuf` and `pop` have a mean significantly larger than the median. For `manuf`, it was found that it has a mean of 463.1 and a median of 347. For `pop`, the results were 608 to 515 respectively.
Furthermore, the maximum values (3344, 3369) are in order of magnitude greater than the 75th percentile (462, 717) for both variables.

With this we could infer that the dataset does not contain 41 similar cities. It contains a majority of "typical" cities and a few massive outliers, metropolitan big cities. In order to avoid that our analysis
was dominated by these few outliers, we decided to use a correlation analysis to normalize our data.

We also noticed that the data is in different units, and every unit measured different things which made
very important the use of the *correlation matrix* to perform this first type of analysis.

Adding up to that, we can check the standard deviation of the variables:

```{r dispersion-measures}
# Calculate standard deviation for each variable.
library(dplyr)

data5_variables %>% summarise_if(is.numeric, sd)
```


Since the standard deviations for `pop` and `manuf` are orders of magnitude larger than for other
variables, we can confirm their high variance.

For example, the sd of `pop` is 579.733 and the sd of `wind` is 1.4. In this case if we run a PCA with this data, it will determine that all the variation is happening around `pop` and not `wind`.
This is why we also need to perform a correlation analysis so that all our variables can have a standard deviation of 1 and a mean of 0.

When we run the PCA on the correlation matrix we will do after this, we can find the directions (PCs) that will maximize the standardized variance based on the correlation relationships between the variables.

For the reason mentioned above and this one, it's better that we use the *correlation matrix* in order to perform our PCA.

## Correlation Analysis + Correlation Matrix

```{r pairs-plot}
# Visualize pairwise relationships between all variables.
pairs(data5_variables, main="Pairs Plot Air Pollution Data")
```

**CORRELATION MATRIX**
```{r correlation-matrix}
# 1st) Calculate and display the correlation matrix.
cor_data5 <- cor(data5_variables)
cor_data5
```


```{r}
library(GGally)

p <- ggpairs(
  data5_variables,
  title = "Pairs Plot : Air Pollution Data",
  upper = list(continuous = wrap("cor", size = 3, color="darkblue")),
  lower = list(continuous = wrap("points", alpha = 0.6, size=1.5, color="darkred")),
  diag = list(continuous = wrap("densityDiag", alpha=0.5, fill="skyblue"))
)
p
ggsave("plots/pairs_plot.png", plot = p, width = 10, height = 8, dpi = 300)

```

By looking at the plot and the correlation values between the variables we can observe that the population size and the number of manufacturing enterprises employing 20 or more workers is strongly correlated (0.955). We can see that the most populated cities in this study (looking at the initial data, data5), employ more people.

Then, we can observe that the number of manufacturing enterprises employing workers and the sulfur dioxide content of air (mg/m3) are correlated as well (0.645), which means that the manufacturing enterprises that employ 20 or more workers end up influencing so2 levels, which makes sense.

Population and so2 levels are correlated as well (0.494), and precipitation and number of days too (0.496).

Even though with a lower value, precipitation  and temperature variables are correlated (0.386).

Summing everything up, the correlation matrix reveals several key relationships:

- A very strong positive correlation (0.955) between `manuf` and `pop`.

- A moderate positive correlation (0.65) between `manuf` and `so2`.

This suggests that industrial and population factors are closely linked and are associated with higher SO2 pollution.



# Principal Component Analysis (PCA)

PCA will be used to reduce the dimensionality of the dataset. By creating a smaller set of
uncorrelated components, we can simplify the data while retaining most of the original variance.
We will use the *correlation matrix* (`scale = TRUE`) to account for the different units and scales of the original variables.

**EIGENVALUES AND EIGENVECTORS**
```{r}
## 2nd) Obtain eigenvectors and eigenvalues

eigen_data5 <- eigen(cor_data5)
eigen_data5
```

## Justification for Number of Components Retained

### Criteria 1 => Kaiser's Criteria

*Analyzing the results*, we should retain the first 3 PCs because there are 3 eigenvalues that are >1 (Kaiser's Criteria).

However, there are other criteria that we should take into consideration. Now we will perform PCA to see if 3 PCs explain the majority of the variance or not.

### Criteria 2 => PCA

```{r pca analysis}
# 3rd) Perfom PCA

pca_results <- princomp(data5_variables, cor = TRUE)

print(summary(pca_results), loadings = TRUE)
```

*Analyzing the results*, we have 81% of the variance explained with the first 3 PCs (viewed in the cumulative proportion). However, if we consider 4 PCs, we will have 93% of the variance explained.

Since this is not conclusive, we should do the third criteria.

### Criteria 3 => Scree Plot

Now we have all that we need to do is a *scree plot* to see where the elbow point might be:

```{r scree-plot}
library(ggplot2)

eigenvalues <- eigen_data5$values

scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)
ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line() +
  geom_point(size=3) +
  ylim(0,1) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()

```

### Criteria 3 => Scree Plot Enhance with Kaizer crteria.

```{r enhanced-screeplot, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)


eigenvalues <- eigen_data5$values
scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)

kaiser_cutoff <- 1 / length(eigenvalues)

ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "darkred") +
  geom_text(aes(label = round(variance_explained, 2)), vjust = -1.2, size = 3) +
  geom_hline(yintercept = kaiser_cutoff, linetype = "dashed", color = "red") +
  labs(
    title = "Scree Plot with Kaiser Criterion",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:length(eigenvalues)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  theme_classic(base_size = 14)
ggsave("plots/scree_plot.png", width = 8, height = 5, dpi = 300)

```

*Analyzing the results*, in this plot, the elbow point would be between PC2 and PC3. However, if we consider the other criterias we have applied before, it would be best to use 3 PCs, having approximately 81% of the variance explained.

**SO LET'S CONSIDER 3 PCs.**

Now, we need to see which are the variables that contribute more for the explanation of each principal component retained.
We need to have a correlation matrix between the variables and the PCs (the loadings).


## Component Interpretation (Loadings)

```{r component matrix}
# Correlation between original variables and the PCs (loadings)

component_matrix <- cor(data5_variables,pca_results$scores)
component_matrix
```

We only need to look into the the 3 first PCs, since we determined we are going to use the first three.

- **Principal Component 1 (PC1):** `manuf`, `pop`, and `so2` explained by high loadings. This is an **"Industrialization and Urbanization"** axis.

- **Principal Component 2 (PC2):** `precip` and `days` explained by high loadings. This is a **"Climate"** axis, separating cold and rainy cities from warm and dry ones.

- **Principal Component 3 (PC3):** `temp` is explained by a high loading.

- `wind` is the only variable that is not well explained by the 3 first PCs. It's only explained well by PC4, which we will not consider.

To verify that we made the right decisions:

# Tables to summarize the loadings

```{r}
library(gridExtra)
library(grid)

# Create table of loadings for the 3 PCs
loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

# Create a visual table.
loadings_plot <- tableGrob(
  loadings_table,
  rows = NULL,
  theme = ttheme_default(
    core = list(fg_params = list(hjust = 0.5, x = 0.5)),
    colhead = list(fg_params = list(fontface = "bold", hjust = 0.5))
  )
)

# # Export as png
if(!dir.exists("./plots")) dir.create("./plots")
png("./plots/pca_loadings_table.png", width = 900, height = 400)
grid.draw(loadings_plot)
# dev.off()

```
# Style 2 of table.
```{r}

library(knitr)

library(kableExtra)

loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

kable(loadings_table, format = "html", digits = 4, align = "c",
      col.names = c("Variable", "PC1", "PC2", "PC3")) %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))

```



Let's prove that with:

```{r}
# Apply the rule for the 1st PC
# lambda1 / p

sqrt(eigen_data5$values[1]/8)
```
We considered everything above this value for PC1.

```{r}
# Apply the rule for the 2nd PC
# lambda1 / p

sqrt(eigen_data5$values[2]/8)
```
We considered everything above this value for PC2.

```{r}
# Apply the rule for the 3rd PC
# lambda1 / p

sqrt(eigen_data5$values[3]/8)
```
We considered everything above this value for PC3.


## Visualization of the Data

```{r scores}
######Scores
#Put the scores in their own variable to use
scores <- as.data.frame(pca_results$scores)

#City names
scores$city <- data5$city
```

The next 3 graphics are on the first 3 PCs:

# Biplots.

## PC1 vs PC2

```{r biplot12, echo=FALSE}
library(ggbiplot)
ggbiplot(pca_results, choices = c(1,2),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC2: Climate (Cold/Wet vs. Warm/Dry)")
```

## PC2 vs PC3

```{r biplot23, echo=FALSE}
library(ggbiplot)
ggbiplot(pca_results, choices = c(2,3),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC2: Climate (Cold/Wet vs. Warm/Dry), PC3: Temperature")
```

## PC1 vs PC3

```{r biplot13, echo=FALSE}
library(ggbiplot)
ggbiplot(pca_results, choices = c(1,3),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC3: Temperature")
```


## Enhance Option 1 :  PC1 vs PC2.
```{r option1, echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.1 <- loadings$Dim.1 * scale_factor
loadings_scaled$Dim.2 <- loadings$Dim.2 * scale_factor

ggplot(scores, aes(x = Dim.1, y = Dim.2)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.1, yend = Dim.2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.1, y = Dim.2, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC2: Climate Gradient",
       x = paste0("PC1 (", round(pca_result$eig[1,2], 2), "%)"),
       y = paste0("PC2 (", round(pca_result$eig[2,2], 2), "%)")) +
  theme_classic(base_size = 14)
ggsave("plots/biplot_pc1_pc2_plot.png", width = 8, height = 5, dpi = 300)


aes(x = Dim.2, y = Dim.3)   # for PC2 vs PC3
aes(x = Dim.1, y = Dim.3)   # for PC1 vs PC3

```

## Option 2.
```{r pca-biplot-ggfortify, message=FALSE, warning=FALSE, echo=FALSE}
library(ggfortify)
autoplot(pca_results,
           data = data5, # Use original data for labels
           loadings = TRUE, loadings.colour = 'red',
           loadings.label = TRUE, loadings.label.size = 3,
           label = TRUE, label.size = 2.5, shape = FALSE) +
    labs(title = "PCA Biplot of US Cities",
         subtitle = "Principal Components and Original Variable Loadings") +
    theme_minimal()
```

## Enhance Option 1 PC2 vs PC3.

## Option 1

```{r pc2-pc3, echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.2 <- loadings$Dim.2 * scale_factor
loadings_scaled$Dim.3 <- loadings$Dim.3 * scale_factor

ggplot(scores, aes(x = Dim.2, y = Dim.3)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.2, yend = Dim.3),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.2, y = Dim.3, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC2: Climate (Cold/Wet vs. Warm/Dry), PC3: Temperature",
       x = paste0("PC2 (", round(pca_result$eig[2,2], 2), "%)"),
       y = paste0("PC3 (", round(pca_result$eig[3,2], 2), "%)")) +
  theme_classic(base_size = 14)

ggsave("plots/biplot_pc2_pc3_plot.png", width = 8, height = 5, dpi = 300)

aes(x = Dim.1, y = Dim.3)   # for PC2 vs PC3
aes(x = Dim.2, y = Dim.3)   # for PC1 vs PC3

```
## Option 2
```{r pca-biplot2, message=FALSE, warning=FALSE, echo=FALSE}
library(ggfortify)
autoplot(pca_results,
           data = data5, # Use original data for labels
           x = 2, y = 3, # <-- specify PC2 vs PC3
           loadings = TRUE, loadings.colour = 'red',
           loadings.label = TRUE, loadings.label.size = 3,
           label = TRUE, label.size = 2.5, shape = FALSE) +
    labs(title = "PCA Biplot of US Cities",
         subtitle = "PC2: Climate (Cold/Wet vs. Warm/Dry), PC3: Temperature") +
    theme_minimal()
```

## Enhance Option 1 PC1 vs PC3.

## Option 1.

```{r pca-biplot3, message=FALSE, warning=FALSE, echo=FALSE}
library(ggfortify)
autoplot(pca_results,
           data = data5, # Use original data for labels
           x = 1, y = 3, # <-- specify PC2 vs PC3
           loadings = TRUE, loadings.colour = 'red',
           loadings.label = TRUE, loadings.label.size = 3,
           label = TRUE, label.size = 2.5, shape = FALSE) +
    labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC3: Temperature") +
    theme_minimal()
```

## OPtion 2.

```{r , echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.1 <- loadings$Dim.1 * scale_factor
loadings_scaled$Dim.3 <- loadings$Dim.3 * scale_factor

ggplot(scores, aes(x = Dim.1, y = Dim.3)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.1, yend = Dim.3),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.1, y = Dim.3, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC3: Temperature",
       x = paste0("PC1 (", round(pca_result$eig[1,2], 2), "%)"),
       y = paste0("PC3 (", round(pca_result$eig[3,2], 2), "%)")) +
  theme_classic(base_size = 14)

ggsave("plots/biplot_pc1_pc3_plot.png", width = 8, height = 5, dpi = 300)


```


## 3D plots.

The biplot provides a visual confirmation of our interpretations and shows how cities relate to these new axes.

So, if we did a graphic for the first 3 PCs:

*This 3D graphic is not visible in the pfd format.*
```{r, eval=TRUE}
library(plotly)

# Supondo que você tenha feito o PCA:
data5_2 <- prcomp(data5_variables, scale. = TRUE)

pca_scores <- as.data.frame(data5_2$x)

raw_standardized_data <- as.data.frame(scale(data5_variables))

pac_socres_data <- pc_scores_data <- as.data.frame(pca_results$scores[, 1:3])

var_explained <- (data5$sdev)^2 / sum((data5$sdev)^2)

pc1_label <- paste0("PC1 (", round(var_explained[1] * 100, 1), "%)")
pc2_label <- paste0("PC2 (", round(var_explained[2] * 100, 1), "%)")
pc3_label <- paste0("PC3 (", round(var_explained[3] * 100, 1), "%)")

plot_ly(pca_scores,
        x = ~PC1,
        y = ~PC2,
        z = ~PC3,
        type = 'scatter3d',
        mode = 'markers',
        text = rownames(pca_scores),
        marker = list(size = 4)
)%>%
  layout(
    title = "3D Graphic with the first 3 PCs",
    scene = list(
      xaxis = list(title = pc1_label),
      yaxis = list(title = pc2_label),
      zaxis = list(title = pc3_label)
    )
  )

```

## Interactive 3D Scatter Plot (Another Version)

A 2D biplot can only show two components at a time. Since our first three components capture over 80% of the variance, an interactive 3D scatter plot is an effective way to visualize the relationships between cities across these dimensions simultaneously. Compare this implementation with your previous attempt to understand how to leverage existing PCA results and correctly label axes.

```{r 3d-pca-plot-v2, message=FALSE, warning=FALSE}
library(plotly)

# 1. Extract the scores and city names from our existing PCA results
scores_df <- as.data.frame(pca_results$scores)
scores_df$city <- data5$city

# 2. Get the proportion of variance explained for axis labels
# This comes from the summary of our pca_results object
pca_summary <- summary(pca_results)
variance_explained <- pca_summary$sdev^2 / sum(pca_summary$sdev^2)

pc1_label <- paste0("PC1 (", round(variance_explained[1] * 100, 1), "%)")
pc2_label <- paste0("PC2 (", round(variance_explained[2] * 100, 1), "%)")
pc3_label <- paste0("PC3 (", round(variance_explained[3] * 100, 1), "%)")

# 3. Create the interactive 3D plot
plot_ly(data = scores_df,
        x = ~Comp.1, y = ~Comp.2, z = ~Comp.3,
        text = ~city, # Use city names for hover text
        type = 'scatter3d',
        mode = 'markers',
        marker = list(size = 5, color = ~Comp.1, colorscale = 'Viridis', showscale = TRUE, colorbar = list(title = pc1_label))) %>%
  layout(
    title = "Interactive 3D PCA Scatter Plot of US Cities (Improved)",
    scene = list(
      xaxis = list(title = pc1_label),
      yaxis = list(title = pc2_label),
      zaxis = list(title = pc3_label)
    )
  )
```


# Cluster Analysis

Now we will do a completely different analysis and group similar cities using *hierarchical clustering*. We will perform this in two forms, with the original data and with the PCA scores, since it's normal to use the latter in some types of analyses.

```{r libraries}
#Load the libraries needed
library(tidyverse) #data manipulation
library(cluster) #clustering algorithms
library(factoextra) #clustering visualization
library(dendextend) #for comparing two dendograms
library(laGP) #squared Euclidean distance
library(metan) #clustering algorithms
library(tidygeocoder) # for geocoding
library(maps) # for map data

```

## Cluster Analysis with the Original Data

*First*, we can view again the correlation matrix between the original variables:

```{r}
cor_data5
```

*Secondly*,  to perform and obtain the *dendograms* we need to determine the distances (to cluster the groups).
```{r, echo=FALSE}
# install.packages("proxy")
library(proxy)
# We Scale the data to perform the euclidean distance.
data5_v <- scale(data5_variables)

#dissimilarity matrix : Mahalanobis distance
# dist_data5 <- proxy::dist(data5_v, method = "Mahalanobis")

# Eucledian distance.
dist_data5 <- dist(data5_v, method = "euclidean")
dist_data5
```
The dissimilarity matrix is calculated from the data matrix and it contains the values measuring the dissimilarity/similarity between the elements of each pair.
Euclidean distance should not be used when we have different variables measured in different units so, in our case we should use Mahalanobis distance, because, besides the different units, we have correlated variables in the correlation matrix, mainly manuf and pop. So, this method would solve our problem with different scales while also the effects of correlation between variables.

To determine the optimal number of clusters in a dendogram for hierarchical methods and for the original data, we used the silhouette width. The best value of k is the one with the largest average silhouette.

## Elbow Method

```{r}
#non hierarchical, n da para usar. Silhoette da para usar.
library(factoextra)
fviz_nbclust(data5_v, hcut, method = "wss")
```
(Here we can see the elbow point at 3 clusters, or even 4 clusters, but we considered the optimal number of clusters k=3)

**Let´s perform the Hierarchical Clustering**.

### SINGLE LINKAGE

```{r}
hc1_s <- agnes(dist_data5, method = "single")
hc1_s$merge
```

```{r}
hc1_s$order
```

## The AC value is very high.

```{r}
hc1_s$ac
```


```{r}
#Dendrogram
dend1_s <- as.dendrogram(hc1_s)
#labels(dend1_s) <- data5$city
order_in_dend <- order.dendrogram(dend1_s)
labels(dend1_s) <- data5$city[order_in_dend]
plot(dend1_s, main = "Single Linkage - standardized data")
abline(h = 3, col = "red", lwd = 2)
```

## Other Dendogram.

```{r}
# assume hc1_s is an agnes object
dend1_s <- as.dendrogram(hc1_s)
dend1_s <- ladderize(dend1_s)                # tidy ordering
labels(dend1_s) <- data5$city[order.dendrogram(dend1_s)]

# plot with nicer defaults
par(mar = c(6, 4, 4, 2) + 0.1)               # room for rotated labels
plot(dend1_s, main = "Single Linkage — Standardized data",
     ylab = "Height (dissimilarity)",        # clearer axis label
     xlab = "",                              # no x label
     sub = "",                               # remove subtitle
     axes = TRUE,                            # show axis
     edgePar = list(col = "darkgray", lwd = 2),
     nodePar = list(cex = 0.7))              # node label size

# rotate/angle labels for readability
# base plot doesn't rotate labels; use text() to add rotated labels if needed
# add cut line and rectangles for k clusters
abline(h = 3, col = "red", lwd = 2, lty = 2)

```


In the single linkage case, it is produced a chain effect, so the elbow method for this case isn't that reliable.
We know that in a dendogram, the most similar objects get merged first and as we go up in the dendogram, less similar objects merge, represented by taller lines.
Merging things that are not similar is not a good cluster, so in this case, the most part of the merges occur below height 2 and at height 3,  the
lines get taller, so, we cut the dendogram from there to decide the optimal number of clusters - in this case, k=3.

```{r}
plot(dend1_s, main="Single Linkage - 3 clusters")
# k =  clusters considered
rect.hclust(hc1_s, k=3, border = 2:4)
```

## Now we can do the *Cophenetic Coefficient*.

```{r}
#Cophenetic Coefficient
clust_single_s <- hclust(dist_data5, method="single")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_1_s <- cophenetic(clust_single_s)

#Cophenetic coefficient
cor(dist_data5, coph_1_s)
```


### COMPLETE LINKAGE


```{r}
hc2_s <- agnes(dist_data5, method = "complete")
hc2_s$merge
```

```{r}
hc2_s$order
```

## AC Value Complete

```{r}
hc2_s$ac
```

The AC value is high.

```{r}
#Dendrogram
dend2_s <- as.dendrogram(hc2_s)
#labels(dend1_s) <- data5$city
order_in_dend2 <- order.dendrogram(dend2_s)
labels(dend2_s) <- data5$city[order_in_dend2]
plot(dend2_s, main = "Complete Linkage - standardized data")
abline(h = 6, col = "red", lwd = 2)
```

## Elbow method for the best linkage method.

```{r Elbow method for the best linkage method}
raw_standardized_data <- data5_v

# Assuming 'pca_scores' are your 3 PC scores
# and 'hc_complete_pca' is your hclust object from Complete Linkage
#
# The fviz_nbclust function can take the original data and apply
# the clustering for you for a range of k.


library(factoextra)
elbow_plot_raw <- fviz_nbclust(raw_standardized_data,
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "wss")

elbow_plot_pca <- fviz_nbclust(pac_socres_data,
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "wss")

 # 2. Extract and combine
 data_elbow_pca <- elbow_plot_pca$data
 data_elbow_raw <- elbow_plot_raw$data

 data_elbow_pca$Approach <- "Robust (PCA)"
 data_elbow_raw$Approach <- "Standarized (Raw)"

 combined_elbow_data <- rbind(data_elbow_pca, data_elbow_raw)

 # 3. Create superior comparison plot
 ggplot(combined_elbow_data, aes(x = clusters, y = y, group = Approach, color = Approach)) +
   geom_line(linewidth = 1.2) +
   geom_point(size = 3) +
  labs(title = "Elbow plot (WSS): Raw vs PCA (complete linkage)",
       subtitle = "Robust (PCA) vs. Standarized (Raw) Approach",
       x = "Number of Clusters (k)",
       y = "Total within-cluster sum of squares (WSS)"
   ) +
   theme_minimal(base_size = 14) +
   scale_color_manual(values = c("Robust (PCA)" = "blue", "Standarized (Raw)" = "red"))

```



## Silhouette for raw data

```{r}

# Silhouette for raw data
fviz_nbclust(raw_standardized_data,
    FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
    method = "silhouette")

```

## Silhouette combine

```{r Silhouette combine}
# 1. Generate data for Silhouette method
 sil_plot_pca <- fviz_nbclust(pac_socres_data, FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"), method = "silhouette")
 sil_plot_raw <- fviz_nbclust(raw_standardized_data, FUNcluster = function(x, k) hcut(x, k, hc_method =  "complete"), method = "silhouette")

 # 2. Extract and combine
 data_sil_pca <- sil_plot_pca$data
 data_sil_raw <- sil_plot_raw$data

 data_sil_pca$Approach <- "Robust (PCA)"
 data_sil_raw$Approach <- "Standarized (Raw)"

 combined_sil_data <- rbind(data_sil_pca, data_sil_raw)

 # 3. Create superior comparison plot
 ggplot(combined_sil_data, aes(x = clusters, y = y, group = Approach, color = Approach)) +
   geom_line(linewidth = 1.2) +
   geom_point(size = 3) +
   labs(
     title = "Comparison of Average Silhouette Scores",
     subtitle = "Robust (PCA) vs. Standarized (Raw) Approach",
     x = "Number of Clusters (k)",
     y = "Average Silhouette Width"
   ) +
   theme_minimal(base_size = 14) +
   scale_color_manual(values = c("Robust (PCA)" = "blue", "Standarized (Raw)" = "red"))
```




```{r}
plot(dend2_s, main="Complete Linkage - 4 clusters")
# k =  clusters considered
rect.hclust(hc2_s, k=3, border = 2:4)
```

## Now we can do the *Cophenetic Coefficient*. complete

```{r}
#Cophenetic Coefficient
clust_complete_s <- hclust(dist_data5, method="complete")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_2_s <- cophenetic(clust_complete_s)

#Cophenetic coefficient
cor(dist_data5, coph_2_s)
```

It's a smaller value than the *Single Linkage* one.

But using the silhouette method to determine the optimal number of clusters:

```{r}
# Define a range of clusters to test
k_values <- 2:6

for (k in k_values) {
  clusters <- cutree(hc2_s, k = k)           # cluster assignments
  sil <- silhouette(clusters, dist_data5)    # silhouette calculation
  avg_sil <- mean(sil[, 3])                  # average silhouette width
  cat("k =", k, "Average silhouette width =", round(avg_sil, 3), "\n")
}
```
The k with highest average silhouette width, k=2, is the optimal number of clusters when using this method.

Considering 2 clusters:
```{r}
plot(dend2_s, main="Complete Linkage - 2 clusters")
# k =  clusters considered
rect.hclust(hc2_s, k=2, border = 2:4)
```



### WARD LINKAGE

```{r}
hc3_s <- agnes(dist_data5, method = "ward")
hc3_s$merge
```

```{r}
hc3_s$order
```

## AC ward

```{r}
hc3_s$ac
```
The AC value is high.

```{r}
#Dendrogram
dend3_s <- as.dendrogram(hc3_s)
#labels(dend1_s) <- data5$city
order_in_dend3 <- order.dendrogram(dend3_s)
labels(dend3_s) <- data5$city[order_in_dend3]
plot(dend3_s, main = "Ward Linkage - standardized data")
```

```{r}
plot(dend3_s, main="Ward Linkage - 5 clusters")
# k =  clusters considered
rect.hclust(hc3_s, k=5, border = 2:4)
```

## Now we can do the *Cophenetic Coefficient*. Ward

```{r}
#Cophenetic Coefficient
clust_ward_s <- hclust(dist_data5, method="ward")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_3_s <- cophenetic(clust_ward_s)

#Cophenetic coefficient
cor(dist_data5, coph_3_s)
```
Using the silhouette method to determine the optimal number of clusters:

```{r}
library(cluster)

# Choose the range of clusters to test
k_values <- 2:6

for (k in k_values) {
  # Cut dendrogram to get cluster assignments
  clusters <- cutree(hc3_s, k = k)

  # Compute silhouette
  sil <- silhouette(clusters, dist_data5)

  # Average silhouette width
  avg_sil <- mean(sil[, 3])
  cat("k =", k, "Average silhouette width =", round(avg_sil, 3), "\n")
}

```

k=6 is the optimal cluster number for the original data.

```{r}
plot(dend3_s, main="Ward Linkage - 6 clusters")
# k =  clusters considered
rect.hclust(hc3_s, k=6, border = 2:4)
```

**RESULT** The ward method has a much lower coefficient than the other ones.

## Cluster combination

```{r dendrograms-display-and-save, echo=TRUE, message=FALSE, warning=FALSE}
# packages
library(cluster)     # agnes
library(dendextend)  # ladderize, set(..., "labels_cex")

data5_v <- scale(data5_variables)

# compute agglomerative clusterings (single, complete, ward)
hc_single   <- agnes(data5_v, method = "single")
hc_complete <- agnes(data5_v, method = "complete")
hc_ward     <- agnes(data5_v, method = "ward")   # agnes "ward"

# convert to hclust for convenience and create dendrograms
hclust_single   <- as.hclust(hc_single)
hclust_complete <- as.hclust(hc_complete)
hclust_ward     <- as.hclust(hc_ward)

dend_single   <- as.dendrogram(hclust_single)   %>% ladderize()
dend_complete <- as.dendrogram(hclust_complete) %>% ladderize()
dend_ward     <- as.dendrogram(hclust_ward)     %>% ladderize()

# set labels in plotting order and reduce label size
labels(dend_single)   <- data5$city[order.dendrogram(dend_single)]
labels(dend_complete) <- data5$city[order.dendrogram(dend_complete)]
labels(dend_ward)     <- data5$city[order.dendrogram(dend_ward)]

# smaller label text for all dendrograms
dend_single   <- set(dend_single,   "labels_cex", 0.55)
dend_complete <- set(dend_complete, "labels_cex", 0.55)
dend_ward     <- set(dend_ward,     "labels_cex", 0.55)

plot_three_dendros <- function(d1, d2, d3) {
  op <- par(mfrow = c(1, 3), mar = c(8, 4, 3, 1) + 0.1)  # extra bottom margin for labels
  plot(d1, main = "Single linkage", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d2, main = "Complete linkage", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d3, main = "Ward linkage", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  par(op)
}

# Display in the knitted report
plot_three_dendros(dend_single, dend_complete, dend_ward)

# 2) Save identical high-resolution PNG to plots/ (re-run the same plotting code inside png())
if (!dir.exists("plots")) dir.create("plots")
png(filename = "plots/dendrograms_comparison.png", width = 2400, height = 1000, res = 300)
plot_three_dendros(dend_single, dend_complete, dend_ward)
dev.off()
```

# Cluster Analysis with the PCA Data

Add cities names´ to the pca_scores table

```{r}
identical(rownames(pca_scores), rownames(data5))
pca_scores2 <- cbind(city = data5$city, pca_scores)
```

Considering 3 principal components for the PCA data :
```{r}
pca_results$sdev
```

```{r}
pca_results$sdev[1:3]
```
In order to perform and obtain the dendograms, we need to determine the distances
(to cluster the groups).

```{r}
pca_cluster <- pca_scores[,1:3] ; pca_cluster
pca_cluster2 <- pca_scores2[,1:4] ; pca_cluster2 #usar com labels das cidades
```


```{r}
#dissimilarity matrix : squared euclidean distance
library(proxy)
dist_pca <- proxy::dist(pca_cluster,method="euclidean")
dist_pca
```


**Let´s perform the Hierarchical Clustering for the PCA data**.

### SINGLE LINKAGE
```{r}
#hierarchical clustering (aglommerative) using Single Linkage
hcpca1 <- agnes(dist_pca,method="single") #agnes Computes agglomerative hierarchical clustering of the dataset.

```

```{r}
hcpca1$merge
```
Visualize the order in which the objects, clustered, appear in the dendrogram.

```{r}
hcpca1$order
```

The **agglomeration coefficient (AC)** describes the strength of the
clustering structure that has been obtained by **Single Linkage**.
However, this coefficient tends to become larger when n increases, so it should
not be used to compare data sets of very different sizes.
The coefficient takes values from 0 to 1 and it is actually the mean of the
normalized lengths at which the clusters are formed.

## AC PCA data single

```{r}
hcpca1$ac
```
As we can see, the ac is high, meaning that the algorithm performed poorly in this case (far from 1).

**Dendogram**

```{r}
dend1_pca <- as.dendrogram(hcpca1)
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
dend1_pca
plot(dend1_pca,main="Single Linkage for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```

# Combination of methods PCA.

## Cophenetic single

**Cophenetic coefficient** (it gives us the relation between the distance matrix in the beginning and the one obtained from the dendogram) :

```{r}
clust_pca <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "single")
clust_pca$cophenetic
```
The cophenetic coefficient has a quite high value.

```{r}
plot(clust_pca,horiz=FALSE,main="Single Linkage method")
```
Considering 4 clusters:
```{r}
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
plot(dend1_pca,main="Single Linkage for PCA data (4 clsuters)")
rect.hclust(hcpca1,k=4,border=2:4)
```

#### Comparing with the Complete linkage method:

```{r}
hcpca2 <- agnes(pca_cluster,method="complete")
hcpca2$merge
```

```{r}
hcpca2$order
```

## AC complete

```{r}
hcpca2$ac
```

```{r}
dend2_pca <- as.dendrogram(hcpca2)
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
dend2_pca
plot(dend2_pca,main="Complete Linkage for PCA data (3 clusters)")
rect.hclust(hcpca2,k=3,border=2:4)
```

## Cophenetic Complete

**Cophenetic coefficient**

```{r}
clust_pca2 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "complete")
clust_pca2$cophenetic
```

```{r}
plot(clust_pca2,horiz=FALSE,main="Complete Linkage method")
```
Considering 4 clusters:
```{r}
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
plot(dend2_pca,main="Complete Linkage for PCA data (4 clsuters)")
rect.hclust(hcpca2,k=4,border=2:4)
```
Finally, considering the number of optimal clusters given by the silhouette method:
```{r}
library(cluster)

# Choose the range of clusters to test
k_values <- 2:6

for (k in k_values) {
  # Cut dendrogram to get cluster assignments
  clusters <- cutree(dend2_pca, k = k)

  # Compute silhouette
  sil <- silhouette(clusters, dist_data5)

  # Average silhouette width
  avg_sil <- mean(sil[, 3])
  cat("k =", k, "Average silhouette width =", round(avg_sil, 3), "\n")
}
```

```{r}
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
plot(dend2_pca,main="Complete Linkage for PCA data - 2 clusters")
rect.hclust(hcpca2,k=2,border=2:4)
```

The single method was better than the complete method.

```{r}
par(mfrow=c(1,2))
plot(dend1_pca)
plot(dend2_pca)
#plot(clust_pca,horiz=FALSE,main="Single Linkage")
#plot(clust_pca2,horiz=FALSE,main="Complete Linkage")
```

# Finally, comparing with the Ward method:

```{r}
hcpca3 <- agnes(pca_cluster,method="ward")
hcpca3$merge
```

```{r}
hcpca3$order
```

## AC Ward

```{r}
hcpca3$ac
```

```{r}
dend3_pca <- as.dendrogram(hcpca3)
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
dend3_pca
plot(dend3_pca,main="Ward method for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```
**Cophenetic coefficient**

## Cophenetic Ward

```{r}
clust_pca3 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "ward.D")
clust_pca3$cophenetic
```

The cophenetic coefficient has a low value with this method of clustering.

```{r}
plot(clust_pca3,horiz=FALSE,main="Ward method")
```
Considering 4 clusters, instead of 3 :

```{r}
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
plot(dend3_pca,main="Ward method for PCA data (4 clusters)")
rect.hclust(hcpca3,k=4,border=2:4)
```
Finally, using the silhouette method:

```{r}
library(cluster)

# Choose the range of clusters to test
k_values <- 2:6

for (k in k_values) {
  # Cut dendrogram to get cluster assignments
  clusters <- cutree(dend3_pca, k = k)

  # Compute silhouette
  sil <- silhouette(clusters, dist_data5)

  # Average silhouette width
  avg_sil <- mean(sil[, 3])
  cat("k =", k, "Average silhouette width =", round(avg_sil, 3), "\n")
}
```
```{r}
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
plot(dend3_pca,main="Ward method for PCA data - 2 clusters")
rect.hclust(hcpca3,k=2,border=2:4)
```

# Summary linkage methods

## Tabla Raw data.

```{r}
library(knitr)
library(kableExtra)
library(dplyr)

# Create table with the values you provided
diagnostics_table <- data.frame(
  Method = c("Single Linkage", "Complete Linkage", "Ward.D2"),
  Agglomerative_Coefficient = c(0.7033724, 0.8376211, 0.8552635),
  Cophenetic_Correlation = c(0.8533611, 0.6873652, 0.4697547)
)

# Format and render as a styled HTML table
diagnostics_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(format = "html",
        col.names = c("Method", "Agglomerative Coefficient (AC)", "Cophenetic Correlation Coefficient"),
        align = c("l", "r", "r"),
        caption = "Clustering diagnostics for PCA-based dendrograms") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))

```


## Gap Statistic for raw data

```{r Gap Statistic for raw data}
library(factoextra)

# Gap Statistic for PCA scores
set.seed(123)  # reproducibility
fviz_nbclust(
  pca_scores2[, -1],   # exclude the city column, keep only numeric PCs
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "gap_stat",
  nboot = 50
) +
  labs(
    title = "Gap Statistic for PCA-based Clustering",
    subtitle = "Hierarchical clustering with complete linkage",
    x = "Number of clusters (k)",
    y = "Gap Statistic"
  ) +
  theme_minimal(base_size = 14)

gap_stat <- clusGap(pca_scores2[, -1],
                    FUN = function(x, k) hcut(x, k, hc_method = "complete"),
                    K.max = 10, B = 50)

fviz_gap_stat(gap_stat)

```



## Tabla PCA.

```{r}

library(dplyr)
library(knitr)
library(kableExtra)

diagnostics_table2 <- data.frame(
  Method = c("Single Linkage", "Complete Linkage", "Ward.D2"),
  Agglomerative_Coefficient = c(0.8118718, 0.9007324, 0.8958177),
  Cophenetic_Correlation = c(0.8529945, 0.7569114, 0.4519069)
)

diagnostics_table2 %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(format = "html",
        col.names = c("Method", "Agglomerative Coefficient (AC)", "Cophenetic Correlation Coefficient"),
        align = c("l", "r", "r"),
        caption = "Clustering diagnostics (updated)") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))

```

# FULL ANALISYS

```{r ward-check-consistent, echo=TRUE, message=FALSE, warning=FALSE}
# Required packages
library(cluster)     # agnes
# library(metan)     # if you used clustering() from metan; not required here

# 0. PCA scores (assume pca_cluster exists: rows = cities, cols = first 3 PCs)
# If not present, extract from pca_scores:
# pca_cluster <- as.data.frame(pca_scores)[, 1:3]

# 1. Create a single distance object to use everywhere
dist_pca <- dist(pca_cluster, method = "euclidean")

# 2. Compute clusterings consistently
# - agnes for single, complete, and ward (gives AC)
hc_agnes_single   <- agnes(pca_cluster, method = "single")
hc_agnes_complete <- agnes(pca_cluster, method = "complete")
hc_agnes_ward     <- agnes(pca_cluster, method = "ward")   # agnes' ward

# - hclust Ward.D2 (recommended Ward variant for hclust)
hc_hclust_wardD2  <- hclust(dist_pca, method = "ward.D2")

# 3. Agglomerative coefficient (AC) is available from agnes objects
ac_single   <- hc_agnes_single$ac
ac_complete <- hc_agnes_complete$ac
ac_ward_agnes <- hc_agnes_ward$ac

# 4. Cophenetic distances and correlations
# Use cophenetic() on hclust objects. For agnes, convert to hclust first.
coph_single_agnes   <- cophenetic(as.hclust(hc_agnes_single))
coph_complete_agnes <- cophenetic(as.hclust(hc_agnes_complete))
coph_ward_agnes     <- cophenetic(as.hclust(hc_agnes_ward))

coph_ward_hclustD2  <- cophenetic(hc_hclust_wardD2)

# Correlations with the original distance
cop_cor_single   <- cor(as.vector(dist_pca), as.vector(coph_single_agnes))
cop_cor_complete <- cor(as.vector(dist_pca), as.vector(coph_complete_agnes))
cop_cor_ward_agnes <- cor(as.vector(dist_pca), as.vector(coph_ward_agnes))
cop_cor_ward_hclustD2 <- cor(as.vector(dist_pca), as.vector(coph_ward_hclustD2))

# 5. Print a tidy summary so you can compare implementations
summary_df <- data.frame(
  Method = c("Single (agnes)", "Complete (agnes)", "Ward (agnes)", "Ward (hclust: ward.D2)"),
  AC = c(ac_single, ac_complete, ac_ward_agnes, NA),
  Cophenetic = c(cop_cor_single, cop_cor_complete, cop_cor_ward_agnes, cop_cor_ward_hclustD2)
)

print(summary_df, digits = 6)

# 6. If you used another function (e.g., clustering(..., clustmethod="ward.D")),
# compare its cophenetic to the above using the same dist_pca:
# Example (if clust_pca3 exists and has $cophenetic):
if (exists("clust_pca3") && !is.null(clust_pca3$cophenetic)) {
  cat("\nclust_pca3$cophenetic (from your package) =\n")
  print(clust_pca3$cophenetic)
  # If it's a single number, compare directly:
  if (length(clust_pca3$cophenetic) == 1) {
    cat("\nComparison: clust_pca3$cophenetic vs cop_cor_ward_hclustD2:\n")
    print(c(clust_pca3 = clust_pca3$cophenetic, ward_hclustD2 = cop_cor_ward_hclustD2))
  }
}

```



Based on the cophenetic coefficient, Single Linkage was the best method. But the Ward method will present better the results (visualization).

Hierarchical clustering → generates clusters.

Cut tree → assign points to clusters.

Silhouette → evaluates “fit” of points in those clusters.

## Combining of PCA dendrograms.

```{r}
# packages
library(cluster)     # agnes
library(dendextend)  # ladderize, set(..., "labels_cex")

# Ensure PCA scores exist and select first 3 PCs (adjust ncomp if needed)
if (!exists("pca_scores")) {
  stop("pca_scores not found. Run PCA and create 'pca_scores' before this chunk.")
}
ncomp <- min(3, ncol(pca_scores))
pca_cluster <- as.data.frame(pca_scores)[, 1:ncomp, drop = FALSE]

# compute agglomerative clusterings (single, complete, ward) on PCA scores
hc_single_pca   <- agnes(pca_cluster, method = "single")
hc_complete_pca <- agnes(pca_cluster, method = "complete")
hc_ward_pca     <- agnes(pca_cluster, method = "ward")   # agnes "ward"

# convert to hclust for convenience and create dendrograms
hclust_single_pca   <- as.hclust(hc_single_pca)
hclust_complete_pca <- as.hclust(hc_complete_pca)
hclust_ward_pca     <- as.hclust(hc_ward_pca)

dend_single_pca   <- as.dendrogram(hclust_single_pca)   %>% ladderize()
dend_complete_pca <- as.dendrogram(hclust_complete_pca) %>% ladderize()
dend_ward_pca     <- as.dendrogram(hclust_ward_pca)     %>% ladderize()

# set labels in plotting order and reduce label size
if (!is.null(data5$city)) {
  labels(dend_single_pca)   <- data5$city[order.dendrogram(dend_single_pca)]
  labels(dend_complete_pca) <- data5$city[order.dendrogram(dend_complete_pca)]
  labels(dend_ward_pca)     <- data5$city[order.dendrogram(dend_ward_pca)]
} else {
  labels(dend_single_pca)   <- rownames(pca_cluster)[order.dendrogram(dend_single_pca)]
  labels(dend_complete_pca) <- rownames(pca_cluster)[order.dendrogram(dend_complete_pca)]
  labels(dend_ward_pca)     <- rownames(pca_cluster)[order.dendrogram(dend_ward_pca)]
}

# smaller label text for all dendrograms
dend_single_pca   <- set(dend_single_pca,   "labels_cex", 0.55)
dend_complete_pca <- set(dend_complete_pca, "labels_cex", 0.55)
dend_ward_pca     <- set(dend_ward_pca,     "labels_cex", 0.55)

plot_three_pca_dendros <- function(d1, d2, d3) {
  op <- par(mfrow = c(1, 3), mar = c(8, 4, 3, 1) + 0.1)  # extra bottom margin for labels
  plot(d1, main = "Single linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d2, main = "Complete linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d3, main = "Ward linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  par(op)
}

# Display in the knitted report
plot_three_pca_dendros(dend_single_pca, dend_complete_pca, dend_ward_pca)

# Save identical high-resolution PNG to plots/ (re-run the same plotting code inside png())
if (!dir.exists("plots")) dir.create("plots")
png(filename = "plots/pca_dendrograms_comparison.png", width = 2400, height = 1000, res = 300)
plot_three_pca_dendros(dend_single_pca, dend_complete_pca, dend_ward_pca)
dev.off()

```



### Determining the Number of Clusters & Characterization


---

Based on the dendrogram, you must decide how many clusters to form and justify it. Then, you will characterize them.

```{r cluster-characterization}
# GOAL: Cut the tree and assign clusters to characterize them.

# 1. Choose a number of clusters (e.g., 4) based on the dendrogram.
num_clusters <- 4
city_clusters <- cutree(hcpca3, k = num_clusters)

# 2. Add the cluster assignments to the original data frame.
data_with_clusters <- data5 %>%
  mutate(cluster = as.factor(city_clusters))

# 3. Calculate the mean of the original variables for each cluster.
# THIS IS THE CRITICAL INTERPRETATION STEP.
cluster_means <- data_with_clusters %>%
  group_by(cluster) %>%
  summarise_at(vars(so2, temp, manuf, pop, wind, precip, days), mean, na.rm = TRUE)

# 4. Print the cluster profiles. You MUST interpret this table in your report.
print("Cluster Profiles (Mean Values):")
print(cluster_means)
```

### Synthesis: Visualizing Clusters on PCA Biplot

Now that the clusters are defined, we visualize them on the PCA biplot. This synthesizes the two analyses, showing how the identified city groupings relate to the primary axes of variation.

```{r pca-biplot-with-clusters, message=FALSE, warning=FALSE}
# install.packages("FactoMineR")
# install.packages("factoextra")
library(FactoMineR)
library(factoextra)

# 1. Run PCA using FactoMineR for its visualization capabilities.
pca_fm_results <- PCA(data5_variables, scale.unit = TRUE, ncp = 5, graph = FALSE)

# 2. Create the biplot, using the 'cluster' column for the habillage.
# The data_with_clusters dataframe now contains our cluster assignments.
fviz_pca_biplot(pca_fm_results,
                repel = TRUE,
                col.var = "darkblue", # Variables color
                col.ind = "#696969",  # Individuals color
                label = "var",
                habillage = data_with_clusters$cluster, # Color individuals by cluster
                addEllipses = TRUE,
                ellipse.type = "confidence",
                legend.title = "Clusters",
                ggtheme = theme_minimal()) +
  labs(title = "PCA Biplot with City Clusters",
       subtitle = "Clusters determined by Ward's method on PC scores")
```

# Spatial Analysis & Comparison

This section elevates the project by mapping the results. We will investigate if the clusters we identified have a geographic pattern.

**NOTE:** The following chunk requires a working internet connection. If the script hangs, ensure `run_spatial_analysis` is `FALSE` in the `setup-spatial-toggle` chunk.

```{r spatial-analysis, eval=run_spatial_analysis, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
# GOAL: Geocode cities and create a map of the US with cities colored by cluster.

print("Attempting to geocode cities. This may take a moment...")

geocoded_data <- tryCatch({
  data_with_clusters %>%
    geocode(city = city, method = 'osm', verbose = FALSE)
}, error = function(e) {
  warning(paste("Geocoding failed with an error:", e$message), immediate. = TRUE)
  return(NULL) # Return NULL on error
})

if (!is.null(geocoded_data) && all(c("lat", "long") %in% colnames(geocoded_data)) && sum(!is.na(geocoded_data$lat)) > 0) {
  print("Geocoding successful. Generating map...")

  us_map <- map_data("state")

  ggplot() +
    geom_polygon(data = us_map, aes(x = long, y = lat, group = group), fill = "grey80", color = "white") +
    geom_point(data = geocoded_data, aes(x = long, y = lat, color = as.factor(cluster)), size = 3, alpha = 0.8) +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(
      title = "Geographic Distribution of City Clusters",
      subtitle = "Cities grouped by pollution and demographic profiles",
      color = "City Cluster"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
} else {
  print("Geocoding failed or returned no valid coordinates. Cannot generate map.")
}
```
## Discussion of Spatial Patterns


# Conclusion

*   What were the main patterns you found in the data? (e.g., "Our analysis identified two primary axes of variation among US cities: one related to industrialization and a second related to climate.")
*   What types of city clusters did you identify? (e.g., "We found four distinct city profiles: ...")
*   Did these clusters show a geographic pattern?
*   What are the practical implications of these findings for a city planner or environmental agency?

# Bibliography



</div>