---
title: "Project Multivariate data Analysis"
author:
  - "Sara Vasques"
  - "Rita Pinto"
  - "Juan Pablo Martinez Aldana"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
#   pdf_document:
#     latex_engine: xelatex
#     toc: true
#     toc_depth: 2
#     number_sections: true
#     fig_caption: true
#     highlight: tango
#     citation_package: biblatex
# bibliography: ../library.bib
# fontsize: 11pt
# geometry: margin=1in
# header-includes:
#  - \setlength{\headheight}{14pt}
#  - \usepackage{xcolor}
#  - \definecolor{teal}{RGB}{0,128,128}
#  - \usepackage{sectsty}
#  - \sectionfont{\color{teal}}
#  - \subsectionfont{\color{teal}}
#  - \usepackage{titlesec}
#  - \titleformat{\section}{\Large\bfseries\color{teal}}{\thesection}{1em}{}
#  - \titleformat{\subsection}{\large\bfseries\color{teal}}{\thesubsection}{0.5em}{}
#  - \usepackage{fancyhdr}
#  - \pagestyle{fancy}
#  - \fancyhead[L]{\color{teal}Importance of biological samples}
#  - \fancyhead[R]{\color{teal}\thepage}
#  - \fancyfoot{}
---
<div style="text-align: justify;">

# Abstract

Air pollution in urban environments is a significant public health concern.
Some evidence indicates that air pollution exposure has way more consequences than the ones that we thought previously. In terms of adverse health impacts like the reduction in life expectancy, and hospital admissions, birth outcomes, and asthma. Nevertheless, these effects exist in both economically developing and developed countries [1].

Classical PCA can be expressed as a projection-based approach, since it's a method that projects high-dimensional data into a lower-dimensional space by finding the principal components (PC) that best represent the original information. With this, we can do a dimensional reduction while preserving the essential structure of the data. [2]


Due to the reasons mentioned before, this project aims to analyze a dataset of 41 US cities to identify underlying patterns and relationships between pollution levels, climatic conditions, and demographic or industrial factors.

The study utilizes multivariate techniques, primarily **Principal Component Analysis (PCA)** and **Cluster Analysis**, to reduce the complexity of the data and group cities with similar profiles.
The objective is to translate complex statistical relationships into clear, interpretable insights.

The dataset contains the following variables:

 **so2**: Sulfur dioxide content of air in micrograms per cubic meter

 **temp**: Average annual temperature in Fahrenheit

 **manuf**: Number of manufacturing enterprises employing 20 or more workers

 **pop**: Population size (1970 census) in thousands

 **wind**: Average wind speed in miles per hour

 **precip**: Average annual precipitation in inches

 **days**: Average number of days with precipitation per year

# Preliminary analysis of the data

Before applying multivariate methods, a preliminary analysis is essential to understand the individual variables and their pairwise relationships.

```{r setup-spatial-toggle, include=FALSE}
# --- ANALYSIS CONTROL ---
# Set this to TRUE if you have a stable internet connection to generate the city map.
# If the script gets stuck on the geocoding step, set this to FALSE and re-run.
run_spatial_analysis <- FALSE
```

```{r data-loading, echo=FALSE}
# The data is loaded from the folder data/data5.csv

#data5 <- read.csv("./data/data_5.csv", header = TRUE)
data5 <- read.csv("./data/data_5.csv", header = TRUE)
head(data5)
```

```{r data-structure, echo=FALSE}
# Display the dimensions and structure of the dataset.
dim(data5)

str(data5)
```

## Numerical Data

```{r variable-selection, echo=FALSE}
# We select only the numeric variables for the analysis. (I modified the selector using seletc_if
# In order to make it more global and we can apply to any data in the future)

library(dplyr)

data5_variables <- select_if(data5, is.numeric)
head(data5_variables)
```

## Descriptive Statistics

```{r localization-measures, echo=FALSE}
# Calculate and display summary statistics for each variable.
summary(data5_variables)
```


We identify that `manuf` and `pop` have a mean significantly larger than the median. For `manuf`, it was found that it has a mean of 463.1 and a median of 347. For `pop`, the results were 608 to 515 respectively.
Furthermore, the maximum values (3344, 3369) are in order of magnitude greater than the 75th percentile (462, 717) for both variables.

With this we could infer that the dataset does not contain 41 similar cities. It contains a majority of "typical" cities and a few massive outliers, metropolitan big cities. In order to avoid that our analysis
was dominated by these few outliers, we decided to use a correlation analysis to normalize our data.

We also noticed that the data is in different units, and every unit measured different things which made
very important the use of the *correlation matrix* to perform this first type of analysis.

Adding up to that, we can check the standard deviation of the variables:

```{r dispersion-measures}
# Calculate standard deviation for each variable.
library(dplyr)

data5_variables %>% summarise_if(is.numeric, sd)
```


Since the standard deviations for `pop` and `manuf` are orders of magnitude larger than for other
variables, we can confirm their high variance.

For example, the sd of `pop` is 579.733 and the sd of `wind` is 1.4. In this case if we run a PCA with this data, it will determine that all the variation is happening around `pop` and not `wind`.
This is why we also need to perform a correlation analysis so that all our variables can have a standard deviation of 1 and a mean of 0.

When we run the PCA on the correlation matrix we will do after this, we can find the directions (PCs) that will maximize the standardized variance based on the correlation relationships between the variables.

For the reason mentioned above and this one, it's better that we use the *correlation matrix* in order to perform our PCA.

## Correlation Analysis + Correlation Matrix

```{r pairs-plot}
# Visualize pairwise relationships between all variables.
pairs(data5_variables, main="Pairs Plot Air Pollution Data")
```

**CORRELATION MATRIX**
```{r correlation-matrix}
# 1st) Calculate and display the correlation matrix.
cor_data5 <- cor(data5_variables)
cor_data5
```


```{r}
library(GGally)

p <- ggpairs(
  data5_variables,
  title = "Pairs Plot : Air Pollution Data",
  upper = list(continuous = wrap("cor", size = 3, color="darkblue")),
  lower = list(continuous = wrap("points", alpha = 0.6, size=1.5, color="darkred")),
  diag = list(continuous = wrap("densityDiag", alpha=0.5, fill="skyblue"))
)

ggsave("plots/pairs_plot.png", plot = p, width = 10, height = 8, dpi = 300)

```

By looking at the plot and the correlation values between the variables we can observe that the population size and the number of manufacturing enterprises employing 20 or more workers is strongly correlated (0.955). We can see that the most populated cities in this study (looking at the initial data, data5), employ more people.

Then, we can observe that the number of manufacturing enterprises employing workers and the sulfur dioxide content of air (mg/m3) are correlated as well (0.645), which means that the manufacturing enterprises that employ 20 or more workers end up influencing so2 levels, which makes sense.

Population and so2 levels are correlated as well (0.494), and precipitation and number of days too (0.496).

Even though with a lower value, precipitation  and temperature variables are correlated (0.386).

Summing everything up, the correlation matrix reveals several key relationships:

- A very strong positive correlation (0.955) between `manuf` and `pop`.

- A moderate positive correlation (0.65) between `manuf` and `so2`.

This suggests that industrial and population factors are closely linked and are associated with higher SO2 pollution.



# Principal Component Analysis (PCA)

PCA will be used to reduce the dimensionality of the dataset. By creating a smaller set of
uncorrelated components, we can simplify the data while retaining most of the original variance.
We will use the *correlation matrix* (`scale = TRUE`) to account for the different units and scales of the original variables.

**EIGENVALUES AND EIGENVECTORS**
```{r}
## 2nd) Obtain eigenvectors and eigenvalues

eigen_data5 <- eigen(cor_data5)
eigen_data5
```

## Justification for Number of Components Retained

### Criteria 1 => Kaiser's Criteria

*Analyzing the results*, we should retain the first 3 PCs because there are 3 eigenvalues that are >1 (Kaiser's Criteria).

However, there are other criteria that we should take into consideration. Now we will perform PCA to see if 3 PCs explain the majority of the variance or not.

### Criteria 2 => PCA

```{r pca analysis}
# 3rd) Perfom PCA

pca_results <- princomp(data5_variables, cor = TRUE)

print(summary(pca_results), loadings = TRUE)
```

*Analyzing the results*, we have 81% of the variance explained with the first 3 PCs (viewed in the cumulative proportion). However, if we consider 4 PCs, we will have 93% of the variance explained.

Since this is not conclusive, we should do the third criteria.

### Criteria 3 => Scree Plot

Now we have all that we need to do is a *scree plot* to see where the elbow point might be:

```{r scree-plot}
library(ggplot2)

eigenvalues <- eigen_data5$values

scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)
ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line() +
  geom_point(size=3) +
  ylim(0,1) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()

```

### Criteria 3 => Scree Plot Enhance with Kaizer crteria.

```{r enhanced-screeplot, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)


eigenvalues <- eigen_data5$values
scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)

kaiser_cutoff <- 1 / length(eigenvalues)

ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "darkred") +
  geom_text(aes(label = round(variance_explained, 2)), vjust = -1.2, size = 3) +
  geom_hline(yintercept = kaiser_cutoff, linetype = "dashed", color = "red") +
  labs(
    title = "Scree Plot with Kaiser Criterion",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:length(eigenvalues)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  theme_classic(base_size = 14)
ggsave("plots/scree_plot.png", width = 8, height = 5, dpi = 300)

```

*Analyzing the results*, in this plot, the elbow point would be between PC2 and PC3. However, if we consider the other criterias we have applied before, it would be best to use 3 PCs, having approximately 81% of the variance explained.

**SO LET'S CONSIDER 3 PCs.**

Now, we need to see which are the variables that contribute more for the explanation of each principal component retained.
We need to have a correlation matrix between the variables and the PCs (the loadings).


## Component Interpretation (Loadings)

```{r component matrix}
# Correlation between original variables and the PCs (loadins)

component_matrix <- cor(data5_variables,pca_results$scores)
component_matrix
```

We only need to look into the the 3 first PCs, since we determined we are going to use the first three.

- **Principal Component 1 (PC1):** `manuf`, `pop`, and `so2` explained by high loadings. This is an **"Industrialization and Urbanization"** axis.

- **Principal Component 2 (PC2):** `precip` and `days` explained by high loadings. This is a **"Climate"** axis, separating cold and rainy cities from warm and dry ones.

- **Principal Component 3 (PC3):** `temp` is explained by a high loading.

- `wind` is the only variable that is not well explained by the 3 first PCs. It's only explained well by PC4, which we will not consider.

To verify that we made the right decisions:

# Tables to summarize the loadings

```{r}
library(gridExtra)
library(grid)

# Create table of loadings for the 3 PCs
loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

# Create a visual table.
loadings_plot <- tableGrob(
  loadings_table,
  rows = NULL,
  theme = ttheme_default(
    core = list(fg_params = list(hjust = 0.5, x = 0.5)),
    colhead = list(fg_params = list(fontface = "bold", hjust = 0.5))
  )
)

# # Export as png
if(!dir.exists("./plots")) dir.create("./plots")
png("./plots/pca_loadings_table.png", width = 900, height = 400)
grid.draw(loadings_plot)
# dev.off()

```
# Style 2 of table.
```{r}

library(knitr)

library(kableExtra)

loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

kable(loadings_table, format = "html", digits = 4, align = "c",
      col.names = c("Variable", "PC1", "PC2", "PC3")) %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))

```



Let's prove that with:

```{r}
# Apply the rule for the 1st PC
# lambda1 / p

sqrt(eigen_data5$values[1]/8)
```
We considered everything above this value for PC1.

```{r}
# Apply the rule for the 2nd PC
# lambda1 / p

sqrt(eigen_data5$values[2]/8)
```
We considered everything above this value for PC2.

```{r}
# Apply the rule for the 3rd PC
# lambda1 / p

sqrt(eigen_data5$values[3]/8)
```
We considered everything above this value for PC3.


## Visualization of the Data

```{r scores}
######Scores
#Put the scores in their own variable to use
scores <- as.data.frame(pca_results$scores)

#City names
scores$city <- data5$city
```

The next 3 graphics are on the first 3 PCs:

### PC1 vs PC2

```{r biplot12}
library(ggbiplot)
ggbiplot(pca_results, choices = c(1,2),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC2: Climate (Cold/Wet vs. Warm/Dry)")
```

### PC2 vs PC3

```{r biplot23}
library(ggbiplot)
ggbiplot(pca_results, choices = c(2,3),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC2: Climate (Cold/Wet vs. Warm/Dry), PC3: Temperature")
```

### PC1 vs PC3

```{r biplot13}
library(ggbiplot)
ggbiplot(pca_results, choices = c(1,3),labels = data5$city, ellipse = TRUE, circle = TRUE) +
  theme_minimal() +
  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC3: Temperature")
```




The biplot provides a visual confirmation of our interpretations and shows how cities relate to these new axes.

So, if we did a graphic for the first 3 PCs:

*This 3D graphic is not visible in the pfd format.*
```{r, eval=TRUE}
library(plotly)

# Supondo que você tenha feito o PCA:
data5_2 <- prcomp(data5_variables, scale. = TRUE)

pca_scores <- as.data.frame(data5_2$x)

var_explained <- (data5$sdev)^2 / sum((data5$sdev)^2)

pc1_label <- paste0("PC1 (", round(var_explained[1] * 100, 1), "%)")
pc2_label <- paste0("PC2 (", round(var_explained[2] * 100, 1), "%)")
pc3_label <- paste0("PC3 (", round(var_explained[3] * 100, 1), "%)")

plot_ly(pca_scores,
        x = ~PC1,
        y = ~PC2,
        z = ~PC3,
        type = 'scatter3d',
        mode = 'markers',
        text = rownames(pca_scores),
        marker = list(size = 4)
)%>%
  layout(
    title = "3D Graphic with the first 3 PCs",
    scene = list(
      xaxis = list(title = pc1_label),
      yaxis = list(title = pc2_label),
      zaxis = list(title = pc3_label)
    )
  )

```


# Cluster Analysis

Now we will do a completely different analysis and group similar cities using *hierarchical clustering*. We will perform this in two forms, with the original data and with the PCA scores, since it's normal to use the latter in some types of analyses.

```{r libraries}
#Load the libraries needed
library(tidyverse) #data manipulation
library(cluster) #clustering algorithms
library(factoextra) #clustering visualization
library(dendextend) #for comparing two dendograms
library(laGP) #squared Euclidean distance
library(metan) #clustering algorithms
library(tidygeocoder) # for geocoding
library(maps) # for map data

```

## Cluster Analysis with the Original Data

*First*, we can view again the correlation matrix between the original variables:

```{r}
cor_data5
```

*Secondly*,  to perform and obtain the *dendograms* we need to determine the distances (to cluster the groups).
```{r}
#remove columns we don't need for the analysis, based on the PCA (in this case, the column "wind")
data5_v <- scale(data5_variables)

#dissimilarity matrix : squared euclidean distance
dist_data5 <- dist(data5_v, method = "euclidean")

dist_data5
```
We used the *Squared Euclidean distance*.

To determine which values are going to be clustered, we need to look at the values that have the lowest distance between them.
Based on the *dissimilarity matrix*, the pair with the lowest distance should be joint first.

**Let´s perform the Hierarchical Clustering**.

### SINGLE LINKAGE

```{r}
hc1_s <- agnes(dist_data5, method = "single")
hc1_s$merge
```

```{r}
hc1_s$order
```

```{r}
hc1_s$ac
```
The AC value is very high.

```{r}
#Dendrogram
dend1_s <- as.dendrogram(hc1_s)
#labels(dend1_s) <- data5$city
order_in_dend <- order.dendrogram(dend1_s)
labels(dend1_s) <- data5$city[order_in_dend]
plot(dend1_s, main = "Single Linkage - standardized data")
```

```{r}
plot(dend1_s, main="Single Linkage - 3 clusters")
# k =  clusters considered
rect.hclust(hc1_s, k=3, border = 2:4)
```

Now we can do the *Cophenetic Coefficient*.

```{r}
#Cophenetic Coefficient
clust_single_s <- hclust(dist_data5, method="single")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_1_s <- cophenetic(clust_single_s)

#Cophenetic coefficient
cor(dist_data5, coph_1_s)
```


### COMPLETE LINKAGE


```{r}
hc2_s <- agnes(dist_data5, method = "complete")
hc2_s$merge
```

```{r}
hc2_s$order
```

```{r}
hc2_s$ac
```
The AC value is high.

```{r}
#Dendrogram
dend2_s <- as.dendrogram(hc2_s)
#labels(dend1_s) <- data5$city
order_in_dend2 <- order.dendrogram(dend2_s)
labels(dend2_s) <- data5$city[order_in_dend2]
plot(dend2_s, main = "Complete Linkage - standardized data")
```

```{r}
plot(dend1_s, main="Complete Linkage - 3 clusters")
# k =  clusters considered
rect.hclust(hc2_s, k=3, border = 2:4)
```

Now we can do the *Cophenetic Coefficient*.

```{r}
#Cophenetic Coefficient
clust_complete_s <- hclust(dist_data5, method="complete")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_2_s <- cophenetic(clust_complete_s)

#Cophenetic coefficient
cor(dist_data5, coph_2_s)
```

It's a smaller valye than the *Single Linkage* one.


### WARD LINKAGE

```{r}
hc3_s <- agnes(dist_data5, method = "ward")
hc3_s$merge
```

```{r}
hc3_s$order
```

```{r}
hc3_s$ac
```
The AC value is high.

```{r}
#Dendrogram
dend3_s <- as.dendrogram(hc3_s)
#labels(dend1_s) <- data5$city
order_in_dend3 <- order.dendrogram(dend3_s)
labels(dend3_s) <- data5$city[order_in_dend3]
plot(dend3_s, main = "Ward Linkage - standardized data")
```

```{r}
plot(dend3_s, main="Ward Linkage - 5 clusters")
# k =  clusters considered
rect.hclust(hc1_s, k=5, border = 2:4)
```

Now we can do the *Cophenetic Coefficient*.

```{r}
#Cophenetic Coefficient
clust_ward_s <- hclust(dist_data5, method="ward")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_3_s <- cophenetic(clust_ward_s)

#Cophenetic coefficient
cor(dist_data5, coph_3_s)
```

**RESULT** The ward method has a much lower coefficient than the other ones.







## Cluster Analysis with the PCA Data

Add cities names´ to the pca_scores table
```{r}
identical(rownames(pca_scores), rownames(data5))
pca_scores2 <- cbind(city = data5$city, pca_scores)
```

Considering 3 principal components for the PCA data :
```{r}
pca_results$sdev
```

```{r}
pca_results$sdev[1:3]
```
In order to perform and obtain the dendograms, we need to determine the distances
(to cluster the groups).

```{r}
pca_cluster <- pca_scores[,1:3] ; pca_cluster
pca_cluster2 <- pca_scores2[,1:4] ; pca_cluster2 #usar com labels das cidades
```


```{r}
#dissimilarity matrix : squared euclidean distance
dist_pca <- dist(pca_cluster,method="euclidean")
dist_pca
```
**Let´s perform the Hierarchical Clustering for the PCA data**.

### SINGLE LINKAGE
```{r}
#hierarchical clustering (aglommerative) using Single Linkage
hcpca1 <- agnes(dist_pca,method="single") #agnes Computes agglomerative hierarchical clustering of the dataset.

```

```{r}
hcpca1$merge
```



Visualize the order in which the objects, clustered, appear in the dendrogram.

```{r}
hcpca1$order
```
The **agglomeration coefficient (AC)** describes the strength of the
clustering structure that has been obtained by **Single Linkage**.
However, this coefficient tends to become larger when n increases, so it should
not be used to compare data sets of very different sizes.
The coefficient takes values from 0 to 1 and it is actually the mean of the
normalized lengths at which the clusters are formed.

```{r}
hcpca1$ac
```
As we can see, the ac is very high, meaning that the algorithm performed poorly in this case (far from 1).

**Dendogram**

```{r}
dend1_pca <- as.dendrogram(hcpca1)
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
dend1_pca
plot(dend1_pca,main="Single Linkage for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```



With another function to perform clustering :

**Cophenetic coefficient** (it gives us the relation between the distance matrix in the beginning and the one obtained from the dendogram) :

```{r}
clust_pca <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "single")
clust_pca$cophenetic
```
The cophenetic coefficient has a quite high value.

```{r}
plot(clust_pca,horiz=FALSE,main="Single Linkage method")
```




Considering 4 clusters:
```{r}
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
plot(dend1_pca,main="Single Linkage for PCA data (4 clsuters)")
rect.hclust(hcpca1,k=4,border=2:4)
```


#### Comparing with the Complete linkage method:

```{r}
hcpca2 <- agnes(pca_cluster,method="complete")
hcpca2$merge
```

```{r}
hcpca2$order
```

```{r}
dend2_pca <- as.dendrogram(hcpca2)
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
dend2_pca
plot(dend2_pca,main="Complete Linkage for PCA data (3 clusters)")
rect.hclust(hcpca2,k=3,border=2:4)
```
**Cophenetic coefficient**

```{r}
clust_pca2 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "complete")
clust_pca2$cophenetic
```

```{r}
plot(clust_pca2,horiz=FALSE,main="Complete Linkage method")
```
Considering 4 clusters:
```{r}
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
plot(dend2_pca,main="Complete Linkage for PCA data (4 clsuters)")
rect.hclust(hcpca2,k=4,border=2:4)
```


The single method was better than the complete method.

```{r}
par(mfrow=c(1,2))
plot(clust_pca,horiz=FALSE,main="Single Linkage")
plot(clust_pca2,horiz=FALSE,main="Complete Linkage")
```

#Finally, comparing with the Ward method:

```{r}
hcpca3 <- agnes(pca_cluster,method="ward")
hcpca3$merge
```

```{r}
hcpca3$order
```

```{r}
dend3_pca <- as.dendrogram(hcpca3)
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
dend3_pca
plot(dend3_pca,main="Ward method for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```
**Cophenetic coefficient**

```{r}
clust_pca3 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "ward.D")
clust_pca3$cophenetic
```
The cophenetic coefficient has a low value with this method of clustering.

```{r}
plot(clust_pca3,horiz=FALSE,main="Ward method")
```
Let´s consider 4 clusters, instead of 3 :

```{r}
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
plot(dend3_pca,main="Ward method for PCA data (4 clusters)")
rect.hclust(hcpca3,k=4,border=2:4)
```
Based on the cophenetic coefficient, Single Linkage was the best method. But the Ward method will present better the results (visualization).






```{r hierarchical-clustering}
# GOAL: Perform hierarchical clustering on the first two PCA scores.
# 1. Extract the first two columns from pca_results$x.
# 2. Calculate the distance matrix using dist().
# 3. Perform clustering using hclust().
# 4. Plot the resulting dendrogram.
#
# Placeholder:
# pca_scores <- as.data.frame(pca_results$x[, 1:2])
# distance_matrix <- dist(pca_scores)
# hclust_results <- hclust(distance_matrix, method = "complete")
# plot(hclust_results, labels = data5$city, main = "Hierarchical Clustering Dendrogram")
```

### Determining the Number of Clusters & Characterization

Based on the dendrogram, you must decide how many clusters to form and justify it. Then, you will characterize them.

```{r cluster-characterization}
# GOAL: Cut the tree and assign clusters to characterize them.

# 1. Choose a number of clusters (e.g., 4) based on the dendrogram.
num_clusters <- 4
city_clusters <- cutree(hcpca3, k = num_clusters)

# 2. Add the cluster assignments to the original data frame.
data_with_clusters <- data5 %>%
  mutate(cluster = city_clusters)

# 3. Calculate the mean of the original variables for each cluster.
# THIS IS THE CRITICAL INTERPRETATION STEP.
cluster_means <- data_with_clusters %>%
  group_by(cluster) %>%
  summarise_at(vars(so2, temp, manuf, pop, wind, precip, days), mean, na.rm = TRUE)

# 4. Print the cluster profiles. You MUST interpret this table in your report.
print("Cluster Profiles (Mean Values):")
print(cluster_means)
```

# Spatial Analysis & Comparison

This section elevates the project by mapping the results. We will investigate if the clusters we identified have a geographic pattern.

**NOTE:** The following chunk requires a working internet connection. If the script hangs, ensure `run_spatial_analysis` is `FALSE` in the `setup-spatial-toggle` chunk.

```{r spatial-analysis, eval=run_spatial_analysis, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
# GOAL: Geocode cities and create a map of the US with cities colored by cluster.

print("Attempting to geocode cities. This may take a moment...")

geocoded_data <- tryCatch({
  data_with_clusters %>%
    geocode(city = city, method = 'osm', verbose = FALSE)
}, error = function(e) {
  warning(paste("Geocoding failed with an error:", e$message), immediate. = TRUE)
  return(NULL) # Return NULL on error
})

if (!is.null(geocoded_data) && all(c("lat", "long") %in% colnames(geocoded_data)) && sum(!is.na(geocoded_data$lat)) > 0) {
  print("Geocoding successful. Generating map...")

  us_map <- map_data("state")

  ggplot() +
    geom_polygon(data = us_map, aes(x = long, y = lat, group = group), fill = "grey80", color = "white") +
    geom_point(data = geocoded_data, aes(x = long, y = lat, color = as.factor(cluster)), size = 3, alpha = 0.8) +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(
      title = "Geographic Distribution of City Clusters",
      subtitle = "Cities grouped by pollution and demographic profiles",
      color = "City Cluster"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
} else {
  print("Geocoding failed or returned no valid coordinates. Cannot generate map.")
}
```
## Discussion of Spatial Patterns


# Conclusion

*   What were the main patterns you found in the data? (e.g., "Our analysis identified two primary axes of variation among US cities: one related to industrialization and a second related to climate.")
*   What types of city clusters did you identify? (e.g., "We found four distinct city profiles: ...")
*   Did these clusters show a geographic pattern?
*   What are the practical implications of these findings for a city planner or environmental agency?

# Bibliography



</div>