---
title: "Project Multivariate data Analysis"
author:
  - "Sara Vasques"
  - "Rita Pinto"
  - "Juan Pablo Martinez Aldana"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE

---
<div style="text-align: justify;">

```{r}
# 2. Create folder if it doesn't exist
out_dir <- file.path("./plots")   # change base path if needed

```


# Abstract

Air pollution in urban environments is a significant public health concern.
Some evidence indicates that air pollution exposure has way more consequences than the ones that we thought previously. In terms of adverse health impacts like the reduction in life expectancy, and hospital admissions, birth outcomes, and asthma. Nevertheless, these effects exist in both economically developing and developed countries [1].

Classical PCA can be expressed as a projection-based approach, since it's a method that projects high-dimensional data into a lower-dimensional space by finding the principal components (PC) that best represent the original information. With this, we can do a dimensional reduction while preserving the essential structure of the data. [2]

Due to the reasons mentioned before, this project aims to analyze a dataset of 41 US cities to identify underlying patterns and relationships between pollution levels, climatic conditions, and demographic or industrial factors.

The study utilizes multivariate techniques, primarily **Principal Component Analysis (PCA)** and **Cluster Analysis**, to reduce the complexity of the data and group cities with similar profiles.
The objective is to translate complex statistical relationships into clear, interpretable insights.

The dataset contains the following variables:

 **so2**: Sulfur dioxide content of air in micrograms per cubic meter

 **temp**: Average annual temperature in Fahrenheit

 **manuf**: Number of manufacturing enterprises employing 20 or more workers

 **pop**: Population size (1970 census) in thousands

 **wind**: Average wind speed in miles per hour

 **precip**: Average annual precipitation in inches

 **days**: Average number of days with precipitation per year

# Preliminary analysis of the data

Before applying multivariate methods, a preliminary analysis is essential to understand the individual variables and their pairwise relationships.

```{r setup-spatial-toggle, include=FALSE}
# --- ANALYSIS CONTROL ---
# Set this to TRUE if you have a stable internet connection to generate the city map.
# If the script gets stuck on the geocoding step, set this to FALSE and re-run.
run_spatial_analysis <- TRUE
```

```{r data-loading, echo=FALSE}
# The data is loaded from the folder data/data5.csv

data5 <- read.csv("./data/data_5.csv", header = TRUE)
head(data5)
```

```{r data-structure, echo=FALSE}
# Display the dimensions and structure of the dataset.
dim(data5)

str(data5)
```

## Numerical Data

```{r variable-selection, echo=FALSE}
# We select only the numeric variables for the analysis. (I modified the selector using seletc_if
# In order to make it more global and we can apply to any data in the future)

library(dplyr)

data5_variables <- select_if(data5, is.numeric)
head(data5_variables)
```

## Descriptive Statistics

```{r}
#correlation heatmap of the numeric variables

# install.packages(c("corrplot", "reshape2"))
library(corrplot)
library(reshape2)

corr_matrix <- cor(data5_variables, use = "complete.obs")  # handles missing values

# Basic correlation heatmap
corrplot(corr_matrix, method = "color",
         addCoef.col = "black",   # show correlation coefficients
         tl.col = "black",        # axis labels color
         number.cex = 0.8,       # font size for coefficients
         tl.cex = 0.8,            # font size for labels
         title = "Correlation Heatmap for the variables",
         mar = c(0,0,1,0))
```


**In relation with the *corr_matrix* table presented above with all the means, medians, etc**

We identify that `manuf` and `pop` have a mean significantly larger than the median. For `manuf`, it was found that it has a mean of 463.1 and a median of 347. For `pop`, the results were 608 to 515 respectively.
Furthermore, the maximum values (3344, 3369) are in order of magnitude greater than the 75th percentile (462, 717) for both variables.

With this we could infer that the dataset does not contain 41 similar cities. It contains a majority of "typical" cities and a few massive outliers, metropolitan big cities. In order to avoid that our analysis
was dominated by these few outliers, we decided to use a correlation analysis to normalize our data.

We also noticed that the data is in different units, and every unit measured different things which made
very important the use of the *correlation matrix* to perform this first type of analysis.

Adding up to that, we can check the standard deviation of the variables:

```{r dispersion-measures}
# Calculate standard deviation for each variable.
library(dplyr)

data5_variables %>% summarise_if(is.numeric, sd)
```


Since the standard deviations for `pop` and `manuf` are orders of magnitude larger than for other
variables, we can confirm their high variance.

For example, the sd of `pop` is 579.733 and the sd of `wind` is 1.4. In this case if we run a PCA with this data, it will determine that all the variation is happening around `pop` and not `wind`.
This is why we also need to perform a correlation analysis so that all our variables can have a standard deviation of 1 and a mean of 0.

When we run the PCA on the correlation matrix we will do after this, we can find the directions (PCs) that will maximize the standardized variance based on the correlation relationships between the variables.

For the reason mentioned above and this one, it's better that we use the *correlation matrix* in order to perform our PCA.

## Correlation Analysis + Correlation Matrix

```{r pairs-plot}
# Visualize pairwise relationships between all variables.
pairs(data5_variables, main="Pairs Plot Air Pollution Data")
```

**CORRELATION MATRIX**
```{r correlation-matrix}
# 1st) Calculate and display the correlation matrix.
cor_data5 <- cor(data5_variables)
cor_data5
```


```{r}
library(GGally)

p <- ggpairs(
  data5_variables,
  title = "Pairs Plot : Air Pollution Data",
  upper = list(continuous = wrap("cor", size = 3, color="darkblue")),
  lower = list(continuous = wrap("points", alpha = 0.6, size=1.5, color="darkred")),
  diag = list(continuous = wrap("densityDiag", alpha=0.5, fill="skyblue"))
)
p
ggsave("plots/pairs_plot.png", plot = p, width = 10, height = 8, dpi = 300)

```

By looking at the plot and the correlation values between the variables we can observe that the population size and the number of manufacturing enterprises employing 20 or more workers is strongly correlated (0.955). We can see that the most populated cities in this study (looking at the initial data, data5), employ more people.

Then, we can observe that the number of manufacturing enterprises employing workers and the sulfur dioxide content of air (mg/m3) are correlated as well (0.645), which means that the manufacturing enterprises that employ 20 or more workers end up influencing so2 levels, which makes sense.

Population and so2 levels are correlated as well (0.494), and precipitation and number of days too (0.496).

Even though with a lower value, precipitation  and temperature variables are correlated (0.386).

Summing everything up, the correlation matrix reveals several key relationships:

- A very strong positive correlation (0.955) between `manuf` and `pop`.

- A moderate positive correlation (0.65) between `manuf` and `so2`.

This suggests that industrial and population factors are closely linked and are associated with higher SO2 pollution.



# Principal Component Analysis (PCA)

PCA will be used to reduce the dimensionality of the dataset. By creating a smaller set of
uncorrelated components, we can simplify the data while retaining most of the original variance.
We will use the *correlation matrix* (`scale = TRUE`) to account for the different units and scales of the original variables.

**EIGENVALUES AND EIGENVECTORS**
```{r}
## 2nd) Obtain eigenvectors and eigenvalues

eigen_data5 <- eigen(cor_data5)
eigen_data5
```

## Justification for Number of Components Retained

### Criteria 1 => Kaiser's Criteria

*Analyzing the results*, we should retain the first 3 PCs because there are 3 eigenvalues that are >1 (Kaiser's Criteria).

However, there are other criteria that we should take into consideration. Now we will perform PCA to see if 3 PCs explain the majority of the variance or not.

### Criteria 2 => PCA

```{r pca analysis}
# 3rd) Perfom PCA
library(stats)
pca_results <- princomp(data5_variables, cor = TRUE)

print(summary(pca_results), loadings = TRUE)
```

*Analyzing the results*, we have 81% of the variance explained with the first 3 PCs (viewed in the cumulative proportion). However, if we consider 4 PCs, we will have 93% of the variance explained.

Since this is not conclusive, we should do the third criteria.

### Criteria 3 => Scree Plot

Now we have all that we need to do is a *scree plot* to see where the elbow point might be:

```{r scree-plot}
library(ggplot2)

eigenvalues <- eigen_data5$values

scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)
ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line() +
  geom_point(size=3) +
  ylim(0,1) +
  labs(title = "Scree Plot", x = "Principal Component", y = "Proportion of Variance Explained") +
  theme_minimal()

```

### Criteria 3 => Scree Plot Enhance with Kaizer crteria.

```{r enhanced-screeplot, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)


eigenvalues <- eigen_data5$values
scree_data <- data.frame(
  component = 1:length(eigenvalues),
  variance_explained = eigenvalues / sum(eigenvalues)
)

kaiser_cutoff <- 1 / length(eigenvalues)

ggplot(scree_data, aes(x = component, y = variance_explained)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(size = 3, color = "darkred") +
  geom_text(aes(label = round(variance_explained, 2)), vjust = -1.2, size = 3) +
  geom_hline(yintercept = kaiser_cutoff, linetype = "dashed", color = "red") +
  labs(
    title = "Scree Plot with Kaiser Criterion",
    x = "Principal Component",
    y = "Proportion of Variance Explained"
  ) +
  scale_x_continuous(breaks = 1:length(eigenvalues)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, 1)) +
  theme_classic(base_size = 14)
ggsave("plots/scree_plot.png", width = 8, height = 5, dpi = 300)

```

*Analyzing the results*, in this plot, the elbow point would be between PC2 and PC3. However, if we consider the other criterias we have applied before, it would be best to use 3 PCs, having approximately 81% of the variance explained.

**SO LET'S CONSIDER 3 PCs.**

Now, we need to see which are the variables that contribute more for the explanation of each principal component retained.
We need to have a correlation matrix between the variables and the PCs (the loadings).


## Component Interpretation (Loadings)

```{r component matrix}
# Correlation between original variables and the PCs (loadings)

component_matrix <- cor(data5_variables,pca_results$scores)
component_matrix
```

We only need to look into the the 3 first PCs, since we determined we are going to use the first three.

- **Principal Component 1 (PC1):** `manuf`, `pop`, and `so2` explained by high loadings. This is an **"Industrialization and Urbanization"** axis.

- **Principal Component 2 (PC2):** `precip` and `days` explained by high loadings. This is a **"Climate"** axis, separating cold and rainy cities from warm and dry ones.

- **Principal Component 3 (PC3):** `temp` is explained by a high loading.

- `wind` is the only variable that is not well explained by the 3 first PCs. It's only explained well by PC4, which we will not consider.

To verify that we made the right decisions:

# Tables to summarize the loadings

```{r}
library(gridExtra)
library(grid)

# Create table of loadings for the 3 PCs
loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

# Create a visual table.
loadings_plot <- tableGrob(
  loadings_table,
  rows = NULL,
  theme = ttheme_default(
    core = list(fg_params = list(hjust = 0.5, x = 0.5)),
    colhead = list(fg_params = list(fontface = "bold", hjust = 0.5))
  )
)

# # Export as png
if(!dir.exists("./plots")) dir.create("./plots")
png("./plots/pca_loadings_table.png", width = 900, height = 400)
grid.draw(loadings_plot)
# dev.off()

```

# Style 2 of table.

```{r}

library(knitr)

library(kableExtra)

loadings_table <- data.frame(
  Variable = c("so2", "temp", "manuf", "pop", "wind", "precip", "days"),
  PC1 = c(0.8088365434, -0.5208984175, 0.8938494595, 0.8053502866, 0.4127189331, 0.0003093839, 0.4297383099),
  PC2 = c(0.10400859, -0.10900423, -0.27778184, -0.34679989, 0.06821718, 0.76968782, 0.83374415),
  PC3 = c(0.01694887, 0.79975860, 0.31553891, 0.40728458, -0.36763244, 0.58113903, -0.12942257)
)

kable(loadings_table, format = "html", digits = 4, align = "c",
      col.names = c("Variable", "PC1", "PC2", "PC3")) %>%
  kable_styling(full_width = FALSE, position = "center", bootstrap_options = c("striped", "hover"))

```



Let's prove that with:

```{r}
# Apply the rule for the 1st PC
# lambda1 / p

sqrt(eigen_data5$values[1]/8)
```
We considered everything above this value for PC1.

```{r}
# Apply the rule for the 2nd PC
# lambda1 / p

sqrt(eigen_data5$values[2]/8)
```
We considered everything above this value for PC2.

```{r}
# Apply the rule for the 3rd PC
# lambda1 / p

sqrt(eigen_data5$values[3]/8)
```
We considered everything above this value for PC3.


## Visualization of the Data

```{r scores}
######Scores
#Put the scores in their own variable to use
scores <- as.data.frame(pca_results$scores)

#City names
scores$city <- data5$city
```


The next 3 graphics are on the first 3 PCs:

# Biplots.

## PC1 vs PC2

```{r option1, echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.1 <- loadings$Dim.1 * scale_factor
loadings_scaled$Dim.2 <- loadings$Dim.2 * scale_factor

ggplot(scores, aes(x = Dim.1, y = Dim.2)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.1, yend = Dim.2),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.1, y = Dim.2, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC2: Climate Gradient",
       x = paste0("PC1 (", round(pca_result$eig[1,2], 2), "%)"),
       y = paste0("PC2 (", round(pca_result$eig[2,2], 2), "%)")) +
  theme_classic(base_size = 14)
ggsave("plots/biplot_pc1_pc2_plot.png", width = 8, height = 5, dpi = 300)


aes(x = Dim.2, y = Dim.3)   # for PC2 vs PC3
aes(x = Dim.1, y = Dim.3)   # for PC1 vs PC3

```

## PC2 vs PC3


```{r pc2-pc3, echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.2 <- loadings$Dim.2 * scale_factor
loadings_scaled$Dim.3 <- loadings$Dim.3 * scale_factor

ggplot(scores, aes(x = Dim.2, y = Dim.3)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.2, yend = Dim.3),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.2, y = Dim.3, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC2: Climate (Cold/Wet vs. Warm/Dry), PC3: Temperature",
       x = paste0("PC2 (", round(pca_result$eig[2,2], 2), "%)"),
       y = paste0("PC3 (", round(pca_result$eig[3,2], 2), "%)")) +
  theme_classic(base_size = 14)

ggsave("plots/biplot_pc2_pc3_plot.png", width = 8, height = 5, dpi = 300)

aes(x = Dim.1, y = Dim.3)   # for PC2 vs PC3
aes(x = Dim.2, y = Dim.3)   # for PC1 vs PC3

```

## PC1 vs PC3


```{r , echo=FALSE}
library(FactoMineR)
library(factoextra)
library(ggplot2)
library(ggrepel)

# Prepare data (exclude city column)
pca_input <- data5[, -which(names(data5) == "city")]

# Run PCA
pca_result <- PCA(pca_input, scale.unit = TRUE, graph = FALSE)

# Scores (coordinates of cities)
scores <- as.data.frame(pca_result$ind$coord)
scores$city <- data5$city

# Loadings (variable contributions)
loadings <- as.data.frame(pca_result$var$coord)
loadings$var <- rownames(loadings)

scale_factor <- 3  # You can adjust this (try 2–5 depending on spread)
loadings_scaled <- loadings
loadings_scaled$Dim.1 <- loadings$Dim.1 * scale_factor
loadings_scaled$Dim.3 <- loadings$Dim.3 * scale_factor

ggplot(scores, aes(x = Dim.1, y = Dim.3)) +
  geom_point(color = "steelblue", size = 3) +
  geom_text_repel(aes(label = city), size = 3, max.overlaps = 30) +

  # Scaled variable vectors
  geom_segment(data = loadings_scaled,
               aes(x = 0, y = 0, xend = Dim.1, yend = Dim.3),
               arrow = arrow(length = unit(0.25, "cm")),
               color = "red", size = 1) +
  geom_text_repel(data = loadings_scaled,
                  aes(x = Dim.1, y = Dim.3, label = var),
                  color = "red", size = 4) +

  labs(title = "PCA Biplot of US Cities",
       subtitle = "PC1: Industrialization/Urbanization, PC3: Temperature",
       x = paste0("PC1 (", round(pca_result$eig[1,2], 2), "%)"),
       y = paste0("PC3 (", round(pca_result$eig[3,2], 2), "%)")) +
  theme_classic(base_size = 14)

ggsave("plots/biplot_pc1_pc3_plot.png", width = 8, height = 5, dpi = 300)


```

# 3D plots.


```{r 3d-pca-plot-v2, message=FALSE, warning=FALSE}
library(plotly)
data5_2 <- prcomp(data5_variables, scale. = TRUE)

raw_standardized_data <- as.data.frame(scale(data5_variables))

pca_scores_data <- pca_scores_data <- as.data.frame(pca_results$scores[, 1:3])

pca_scores <- as.data.frame(data5_2$x)

# 1. Extract the scores and city names from our existing PCA results
scores_df <- as.data.frame(pca_results$scores)
scores_df$city <- data5$city

# 2. Get the proportion of variance explained for axis labels
# This comes from the summary of our pca_results object
pca_summary <- summary(pca_results)
variance_explained <- pca_summary$sdev^2 / sum(pca_summary$sdev^2)

pc1_label <- paste0("PC1 (", round(variance_explained[1] * 100, 1), "%)")
pc2_label <- paste0("PC2 (", round(variance_explained[2] * 100, 1), "%)")
pc3_label <- paste0("PC3 (", round(variance_explained[3] * 100, 1), "%)")

# 3. Create the interactive 3D plot
plot_ly(data = scores_df,
        x = ~Comp.1, y = ~Comp.2, z = ~Comp.3,
        text = ~city, # Use city names for hover text
        type = 'scatter3d',
        mode = 'markers',
        marker = list(size = 5, color = ~Comp.1, colorscale = 'Viridis', showscale = TRUE, colorbar = list(title = pc1_label))) %>%
  layout(
    title = "Interactive 3D PCA Scatter Plot of US Cities (Improved)",
    scene = list(
      xaxis = list(title = pc1_label),
      yaxis = list(title = pc2_label),
      zaxis = list(title = pc3_label)
    )
  )
```

# Cluster Analysis

Now we will do a completely different analysis and group similar cities using *hierarchical clustering*. We will perform this in two forms, with the original data and with the PCA scores, since it's normal to use the latter in some types of analyses.

```{r libraries}
#Load the libraries needed
library(tidyverse) #data manipulation
library(cluster) #clustering algorithms
library(factoextra) #clustering visualization
library(dendextend) #for comparing two dendograms
library(laGP) #squared Euclidean distance
library(metan) #clustering algorithms
library(tidygeocoder) # for geocoding
library(maps) # for map data

```

## Cluster Analysis with the Original Data

To perform and obtain the *dendograms* we need to determine
the distances (to cluster the groups).

```{r, echo=FALSE}
# We Scale the data to perform the euclidean distance.
data5_v <- scale(data5_variables)

# Eucledian distance.
dist_data5 <- dist(data5_v, method = "euclidean")
dist_data5
```

## SINGLE LINKAGE

```{r}
hc1_s <- agnes(dist_data5, method = "single")
hc1_s$merge
hc1_s$order
```

### AC value.

```{r}
hc1_s$ac
```

### Cophenetic Coefficient.

```{r}
#Cophenetic Coefficient
clust_single_s <- hclust(dist_data5, method="single")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_1_s <- cophenetic(clust_single_s)

#Cophenetic coefficient
cor(dist_data5, coph_1_s)
```

```{r}
#Dendrogram
dend1_s <- as.dendrogram(hc1_s)
#labels(dend1_s) <- data5$city
order_in_dend <- order.dendrogram(dend1_s)
labels(dend1_s) <- data5$city[order_in_dend]
plot(dend1_s, main = "Single Linkage - standardized data")
abline(h = 3, col = "red", lwd = 2)
```

## COMPLETE LINKAGE


```{r}
hc2_s <- agnes(dist_data5, method = "complete")
hc2_s$merge
hc2_s$order
```

### AC Value Complete

```{r}
hc2_s$ac
```

### Cophenetic Coefficient

```{r}
#Cophenetic Coefficient
clust_complete_s <- hclust(dist_data5, method="complete")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_2_s <- cophenetic(clust_complete_s)

#Cophenetic coefficient
cor(dist_data5, coph_2_s)
```

It's a smaller value than the *Single Linkage* one.

### Dendrogram.

```{r}
#Dendrogram
dend2_s <- as.dendrogram(hc2_s)
#labels(dend1_s) <- data5$city
order_in_dend2 <- order.dendrogram(dend2_s)
labels(dend2_s) <- data5$city[order_in_dend2]
plot(dend2_s, main = "Complete Linkage - standardized data")
abline(h = 6, col = "red", lwd = 2)
```


## WARD LINKAGE

```{r}
hc3_s <- agnes(dist_data5, method = "ward")
hc3_s$merge
hc3_s$order
```

### AC ward

```{r}
hc3_s$ac
```
The AC value is high.

### Cophenetic Coefficient*. Ward

```{r}
#Cophenetic Coefficient
clust_ward_s <- hclust(dist_data5, method="ward")

#Cophenetic matrix (delta matrix - obtained based on the dendogram)
coph_3_s <- cophenetic(clust_ward_s)

#Cophenetic coefficient
cor(dist_data5, coph_3_s)
```

### Dendrogram

```{r}
dend3_s <- as.dendrogram(hc3_s)
#labels(dend1_s) <- data5$city
order_in_dend3 <- order.dendrogram(dend3_s)
labels(dend3_s) <- data5$city[order_in_dend3]
plot(dend3_s, main = "Ward Linkage - standardized data")
```

# Cluster Analysis with the PCA Data

Add cities names´ to the pca_scores table

```{r}
identical(rownames(pca_scores), rownames(data5))
pca_scores2 <- cbind(city = data5$city, pca_scores)
#Considering 3 principal components for the PCA data :
pca_results$sdev

pca_results$sdev[1:3]
#In order to perform and obtain the dendograms, we need to determine the distances
#(to cluster the groups).

pca_cluster <- pca_scores[,1:3] ; pca_cluster
pca_cluster2 <- pca_scores2[,1:4] ; pca_cluster2 #usar com labels das cidades

dist_pca <- dist(pca_cluster,method="euclidean")
dist_pca
```

## SINGLE LINKAGE
```{r}
#hierarchical clustering (aglommerative) using Single Linkage
hcpca1 <- agnes(dist_pca,method="single") #agnes Computes agglomerative hierarchical clustering of the dataset.
hcpca1$merge
#Visualize the order in which the objects, clustered, appear in the dendrogram.
hcpca1$order

```

### AC

```{r}
hcpca1$ac
```

### Cophenetic single

```{r}
clust_pca <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "single")
clust_pca$cophenetic
```

### Dendogram

```{r}
dend1_pca <- as.dendrogram(hcpca1)
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
dend1_pca
plot(dend1_pca,main="Single Linkage for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```


## COMPLETE LINKAGE

```{r}
hcpca2 <- agnes(pca_cluster,method="complete")
hcpca2$merge
hcpca2$order
```

### AC complete

```{r}
hcpca2$ac
```

### Cophenetic Coefficient

```{r}
clust_pca2 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "complete")
clust_pca2$cophenetic
```

### Dendrogram


```{r}
dend2_pca <- as.dendrogram(hcpca2)
order_in_dend <- order.dendrogram(dend2_pca)
labels(dend2_pca) <- data5$city[order_in_dend]
dend2_pca
plot(dend2_pca,main="Complete Linkage for PCA data (3 clusters)")
rect.hclust(hcpca2,k=3,border=2:4)
```

## Ward method:


```{r}
hcpca3 <- agnes(pca_cluster,method="ward")
hcpca3$merge
hcpca3$order
```

### AC Ward.

```{r}
hcpca3$ac
```


### Cophenetic Ward.

```{r}
clust_pca3 <- clustering(pca_cluster,distmethod = "euclidean",clustmethod = "ward.D")
clust_pca3$cophenetic
```

### Dendrograma.

```{r}
dend3_pca <- as.dendrogram(hcpca3)
order_in_dend <- order.dendrogram(dend3_pca)
labels(dend3_pca) <- data5$city[order_in_dend]
dend3_pca
plot(dend3_pca,main="Ward method for PCA data (3 clusters)")
rect.hclust(hcpca1,k=3,border=2:4)
```


# Summary linkage methods

## Tabla Raw data.

```{r}
library(knitr)
library(kableExtra)
library(dplyr)

# Create table with the values you provided
diagnostics_table <- data.frame(
  Method = c("Single Linkage", "Complete Linkage", "Ward.D2"),
  Agglomerative_Coefficient = c(0.7033724, 0.8376211, 0.8552635),
  Cophenetic_Correlation = c(0.8533611, 0.6873652, 0.4697547)
)

# Format and render as a styled HTML table
diagnostics_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(format = "html",
        col.names = c("Method", "Agglomerative Coefficient (AC)", "Cophenetic Correlation Coefficient"),
        align = c("l", "r", "r"),
        caption = "Clustering diagnostics for PCA-based dendrograms") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))

```

## Tabla PCA.

```{r}

library(dplyr)
library(knitr)
library(kableExtra)

diagnostics_table2 <- data.frame(
  Method = c("Single Linkage", "Complete Linkage", "Ward.D2"),
  Agglomerative_Coefficient = c(0.8118718, 0.9007324, 0.8958177),
  Cophenetic_Correlation = c(0.8529945, 0.7569114, 0.4519069)
)

diagnostics_table2 %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(format = "html",
        col.names = c("Method", "Agglomerative Coefficient (AC)", "Cophenetic Correlation Coefficient"),
        align = c("l", "r", "r"),
        caption = "Clustering diagnostics (updated)") %>%
  kable_styling(full_width = FALSE, position = "center",
                bootstrap_options = c("striped", "hover"))

```


## Combining of PCA dendrograms.

```{r}
# packages
library(cluster)     # agnes
library(dendextend)  # ladderize, set(..., "labels_cex")

# Ensure PCA scores exist and select first 3 PCs (adjust ncomp if needed)
if (!exists("pca_scores")) {
  stop("pca_scores not found. Run PCA and create 'pca_scores' before this chunk.")
}
ncomp <- min(3, ncol(pca_scores))
pca_cluster <- as.data.frame(pca_scores)[, 1:ncomp, drop = FALSE]

# compute agglomerative clusterings (single, complete, ward) on PCA scores
hc_single_pca   <- agnes(pca_cluster, method = "single")
hc_complete_pca <- agnes(pca_cluster, method = "complete")
hc_ward_pca     <- agnes(pca_cluster, method = "ward")   # agnes "ward"

# convert to hclust for convenience and create dendrograms
hclust_single_pca   <- as.hclust(hc_single_pca)
hclust_complete_pca <- as.hclust(hc_complete_pca)
hclust_ward_pca     <- as.hclust(hc_ward_pca)

dend_single_pca   <- as.dendrogram(hclust_single_pca)   %>% ladderize()
dend_complete_pca <- as.dendrogram(hclust_complete_pca) %>% ladderize()
dend_ward_pca     <- as.dendrogram(hclust_ward_pca)     %>% ladderize()

# set labels in plotting order and reduce label size
if (!is.null(data5$city)) {
  labels(dend_single_pca)   <- data5$city[order.dendrogram(dend_single_pca)]
  labels(dend_complete_pca) <- data5$city[order.dendrogram(dend_complete_pca)]
  labels(dend_ward_pca)     <- data5$city[order.dendrogram(dend_ward_pca)]
} else {
  labels(dend_single_pca)   <- rownames(pca_cluster)[order.dendrogram(dend_single_pca)]
  labels(dend_complete_pca) <- rownames(pca_cluster)[order.dendrogram(dend_complete_pca)]
  labels(dend_ward_pca)     <- rownames(pca_cluster)[order.dendrogram(dend_ward_pca)]
}

# smaller label text for all dendrograms
dend_single_pca   <- set(dend_single_pca,   "labels_cex", 0.55)
dend_complete_pca <- set(dend_complete_pca, "labels_cex", 0.55)
dend_ward_pca     <- set(dend_ward_pca,     "labels_cex", 0.55)

plot_three_pca_dendros <- function(d1, d2, d3) {
  op <- par(mfrow = c(1, 3), mar = c(8, 4, 3, 1) + 0.1)  # extra bottom margin for labels
  plot(d1, main = "Single linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d2, main = "Complete linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  plot(d3, main = "Ward linkage (PCA scores)", ylab = "Height",
       edgePar = list(col = "darkgray", lwd = 1))
  par(op)
}

# Display in the knitted report
plot_three_pca_dendros(dend_single_pca, dend_complete_pca, dend_ward_pca)

# Save identical high-resolution PNG to plots/ (re-run the same plotting code inside png())
if (!dir.exists("plots")) dir.create("plots")
png(filename = "plots/pca_dendrograms_comparison.png", width = 2400, height = 1000, res = 300)
plot_three_pca_dendros(dend_single_pca, dend_complete_pca, dend_ward_pca)
dev.off()

```

# Number clusters.

## Elbow Method

```{r}
#non hierarchical, n da para usar. Silhoette da para usar.
library(factoextra)
fviz_nbclust(data5_v, hcut, method = "wss")
```
(Here we can see the elbow point at 3 clusters, or even 4 clusters, but we considered
the optimal number of clusters k=3)


## Elbow method for the best linkage method.

```{r Elbow method for the best linkage method}
raw_standardized_data <- data5_v

# Assuming 'pca_scores' are your 3 PC scores
# and 'hc_complete_pca' is your hclust object from Complete Linkage
#
# The fviz_nbclust function can take the original data and apply
# the clustering for you for a range of k.


library(factoextra)
elbow_plot_raw <- fviz_nbclust(raw_standardized_data,
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "wss")

elbow_plot_pca <- fviz_nbclust(pca_scores_data,
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "wss")

 # 2. Extract and combine
 data_elbow_pca <- elbow_plot_pca$data
 data_elbow_raw <- elbow_plot_raw$data

 data_elbow_pca$Approach <- "Robust (PCA)"
 data_elbow_raw$Approach <- "Standarized (Raw)"

 combined_elbow_data <- rbind(data_elbow_pca, data_elbow_raw)

 # 3. Create superior comparison plot
 ggplot(combined_elbow_data, aes(x = clusters, y = y, group = Approach, color = Approach)) +
   geom_line(linewidth = 1.2) +
   geom_point(size = 3) +
  labs(title = "Elbow plot (WSS): Raw vs PCA (complete linkage)",
       subtitle = "Robust (PCA) vs. Standarized (Raw) Approach",
       x = "Number of Clusters (k)",
       y = "Total within-cluster sum of squares (WSS)"
   ) +
   theme_minimal(base_size = 14) +
   scale_color_manual(values = c("Robust (PCA)" = "blue", "Standarized (Raw)" = "red"))

```


```{r}
plot(dend1_s, main="Single Linkage - 3 clusters")
# k =  clusters considered
rect.hclust(hc1_s, k=3, border = 2:4)
```

## Silhouette for raw data

```{r}

# Silhouette for raw data
fviz_nbclust(data5_v,
    FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
    method = "silhouette")

```

## Silhouette combine

```{r Silhouette combine}
# 1. Generate data for Silhouette method
 sil_plot_pca <- fviz_nbclust(pca_scores_data, FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"), method = "silhouette")
 sil_plot_raw <- fviz_nbclust(raw_standardized_data, FUNcluster = function(x, k) hcut(x, k, hc_method =  "complete"), method = "silhouette")

 # 2. Extract and combine
 data_sil_pca <- sil_plot_pca$data
 data_sil_raw <- sil_plot_raw$data

 data_sil_pca$Approach <- "Robust (PCA)"
 data_sil_raw$Approach <- "Standarized (Raw)"

 combined_sil_data <- rbind(data_sil_pca, data_sil_raw)

 # 3. Create superior comparison plot
 ggplot(combined_sil_data, aes(x = clusters, y = y, group = Approach, color = Approach)) +
   geom_line(linewidth = 1.2) +
   geom_point(size = 3) +
   labs(
     title = "Comparison of Average Silhouette Scores",
     subtitle = "Robust (PCA) vs. Standarized (Raw) Approach",
     x = "Number of Clusters (k)",
     y = "Average Silhouette Width"
   ) +
   theme_minimal(base_size = 14) +
   scale_color_manual(values = c("Robust (PCA)" = "blue", "Standarized (Raw)" = "red"))
```



# number k=3 COMPLETE LINKAGE

## Gap Statistic for raw data

```{r Gap Statistic for raw data}
library(factoextra)

# Gap Statistic for PCA scores
set.seed(123)  # reproducibility
fviz_nbclust(
  pca_scores2[, -1],   # exclude the city column, keep only numeric PCs
  FUNcluster = function(x, k) hcut(x, k, hc_method = "complete"),
  method = "gap_stat",
  nboot = 50
) +
  labs(
    title = "Gap Statistic for PCA-based Clustering",
    subtitle = "Hierarchical clustering with complete linkage",
    x = "Number of clusters (k)",
    y = "Gap Statistic"
  ) +
  theme_minimal(base_size = 14)

gap_stat <- clusGap(data5_v[, -1],
                    FUN = function(x, k) hcut(x, k, hc_method = "complete"),
                    K.max = 10, B = 50)

fviz_gap_stat(gap_stat)

```

```{r}
plot(dend2_s, main="Complete Linkage - 4 clusters")
# k =  clusters considered
rect.hclust(hc2_s, k=3, border = 2:4)
```

# Combination of methods PCA.



```{r}
plot(clust_pca,horiz=FALSE,main="Single Linkage method")
```

Considering 4 clusters:

```{r}
order_in_dend <- order.dendrogram(dend1_pca)
labels(dend1_pca) <- data5$city[order_in_dend]
plot(dend1_pca,main="Single Linkage for PCA data (4 clsuters)")
rect.hclust(hcpca1,k=4,border=2:4)
```



# Spatial Analysis & Comparison

This section elevates the project by mapping the results. We will investigate if the clusters we identified have a geographic pattern.

**NOTE:** The following chunk requires a working internet connection. If the script hangs, ensure `run_spatial_analysis` is `FALSE` in the `setup-spatial-toggle` chunk.

```{r spatial-analysis, eval=run_spatial_analysis, fig.width=10, fig.height=7, warning=FALSE, message=FALSE}
# GOAL: Geocode cities and create a map of the US with cities colored by cluster.

print("Attempting to geocode cities. This may take a moment...")

geocoded_data <- tryCatch({
  data_with_clusters %>%
    geocode(city = city, method = 'osm', verbose = FALSE)
}, error = function(e) {
  warning(paste("Geocoding failed with an error:", e$message), immediate. = TRUE)
  return(NULL) # Return NULL on error
})

if (!is.null(geocoded_data) && all(c("lat", "long") %in% colnames(geocoded_data)) && sum(!is.na(geocoded_data$lat)) > 0) {
  print("Geocoding successful. Generating map...")

  us_map <- map_data("state")

  ggplot() +
    geom_polygon(data = us_map, aes(x = long, y = lat, group = group), fill = "grey80", color = "white") +
    geom_point(data = geocoded_data, aes(x = long, y = lat, color = as.factor(cluster)), size = 3, alpha = 0.8) +
    coord_map("albers", lat0 = 39, lat1 = 45) +
    labs(
      title = "Geographic Distribution of City Clusters",
      subtitle = "Cities grouped by pollution and demographic profiles",
      color = "City Cluster"
    ) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank()
    )
} else {
  print("Geocoding failed or returned no valid coordinates. Cannot generate map.")
}
```
## Discussion of Spatial Patterns